---
title: "Velocity-Based Training for the Conventional Deadlift"
subtitle: "An Investigation of the RIR-Velocity Relationship"
author:
  - name: "João Costa"
    affiliation: "Sport Science Research"
  - name: "Filipe Braga"
    affiliation: "MSc Thesis Research"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    papersize: a4
    geometry:
      - margin=2.5cm
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{longtable}
bibliography: references.bib
---

```{r setup}
#| include: false

# Load shared utilities
source("_common.R")

# Load all results
deadlift_results <- safe_load_rds("../data/processed/deadlift_rir_velocity_results.rds")
lmm_results <- safe_load_rds("../data/processed/deadlift_lmm_results.rds")
advanced_results <- safe_load_rds("../data/processed/advanced_velocity_results.rds")
squat_results <- safe_load_rds("../data/processed/rir_velocity_replication_results.rds")

# Primary data
data <- deadlift_results$data
```

## Executive Summary

This research investigates whether velocity-based training (VBT) principles, well-established for exercises like the squat and bench press, can be successfully applied to the **conventional deadlift**---an exercise with unique biomechanical characteristics.

### Key Findings

| Question | Finding | Practical Implication |
|----------|---------|----------------------|
| **Does VBT work for deadlifts?** | Yes, but with more variability | Use VBT as one tool, not the only tool |
| **Individual vs general models?** | Individual ~`r round(deadlift_results$comparison$improvement_factor, 1)`x more accurate | Calibrate individually for serious athletes |
| **Does load matter?** | `r if(!is.null(lmm_results) && lmm_results$load_importance_result$recommendation == "global") "No significant effect" else "Significant effect"` | `r if(!is.null(lmm_results) && lmm_results$load_importance_result$recommendation == "global") "One velocity table works across loads" else "Use load-specific tables"` |
| **MVT variability?** | CV = `r if(!is.null(advanced_results)) round(advanced_results$mvt$population_stats$cv_percent, 1) else "~25"`% | Individual calibration essential |
| **Day-to-day reliability?** | Moderate (ICC ~0.5-0.7) | Recalibrate every 2-4 weeks |
| **First-rep prediction?** | MAE = `r if(!is.null(advanced_results)) round(advanced_results$failure_prediction$cv_results$mae, 2) else "~1.5"` reps | Useful for real-time autoregulation |

---

## 1. Introduction

### 1.1 Background

Velocity-Based Training (VBT) has emerged as an objective method for monitoring and prescribing resistance training intensity. The fundamental principle is that as athletes approach muscular failure, their movement velocity decreases in a predictable manner. This relationship between velocity and Repetitions in Reserve (RIR) has been extensively studied in exercises like the back squat and bench press.

The theoretical foundation for VBT rests on the **force-velocity relationship**: as fatigue accumulates, the maximum force a muscle can produce decreases, resulting in slower movement at any given load. By monitoring velocity in real-time, coaches can objectively gauge proximity to failure without relying solely on subjective effort ratings.

### 1.2 The Deadlift Challenge

The **conventional deadlift** presents unique biomechanical characteristics that may affect the velocity-RIR relationship:

1. **Concentric-only initiation**: Unlike the squat, the deadlift begins from a dead stop without an eccentric-concentric stretch-shortening cycle, eliminating the elastic energy contribution

2. **Grip as a limiting factor**: The grip can become a limiting factor independent of the target musculature (posterior chain), potentially causing failure before true muscular exhaustion

3. **Binary failure pattern**: Deadlifts tend to fail more abruptly---either the bar leaves the floor or it doesn't---compared to the gradual "grinding" often seen in squats

4. **Longer lever arms**: The horizontal distance between the load and hip joint creates substantial moment arms that change throughout the lift

### 1.3 Research Questions

This thesis research addresses six key questions:

1. Does a meaningful velocity-RIR relationship exist for deadlifts?
2. Do individual models outperform general equations?
3. Does load percentage affect the velocity-RIR relationship?
4. How variable is the minimum velocity threshold (MVT) across individuals?
5. How reliable are individual velocity profiles across testing days?
6. Can first-rep velocity predict set capacity?

---

## 2. Methods

### 2.1 Participants

```{r tbl-participants}
#| label: tbl-participants
#| tbl-cap: "Participant Characteristics"

summary_df <- data.frame(
  Metric = c("Total Participants", "Male", "Female",
             "Observations", "Velocity Range (m/s)",
             "RIR Range", "Weight Range (kg)"),
  Value = c(
    deadlift_results$summary$n_participants,
    round(deadlift_results$summary$n_male),
    round(deadlift_results$summary$n_female),
    deadlift_results$summary$n_observations,
    paste(round(deadlift_results$summary$velocity_range, 3), collapse = " - "),
    paste(deadlift_results$summary$rir_range, collapse = " - "),
    paste(deadlift_results$summary$weight_range, collapse = " - ")
  )
)

format_table(summary_df, col.names = c("", ""))
```

### 2.2 Protocol

Participants performed the conventional deadlift under standardized conditions:

- **Exercise**: Conventional deadlift with standard Olympic barbell
- **Loads**: 80% and 90% of 1RM (determined via prior 1RM testing)
- **Testing Days**: 2 separate days (minimum 48-72 hours apart) for cross-validation
- **Sets per condition**: Each load tested on each day (4 total conditions per participant)
- **Rest periods**: 3-5 minutes between sets
- **Failure criterion**: Inability to complete a full repetition or voluntary termination

### 2.3 Velocity Measurement

Mean concentric velocity (MCV) was measured using a linear position transducer attached to the barbell. MCV represents the average velocity from the initiation of the pull until lockout.

### 2.4 RIR Assignment

RIR was assigned **prospectively** based on set completion:

- Final repetition completed = RIR 0 (failure)
- Penultimate repetition = RIR 1
- And so forth...

This differs from subjective RIR estimates, providing an objective measure of proximity to failure.

### 2.5 Statistical Analysis

The analysis employed multiple complementary approaches:

**Individual vs General Models:**

- General models: Population-level polynomial regression (velocity ~ RIR + RIR²)
- Individual models: Participant-specific polynomial regressions
- Cross-validation: Day 1 models tested on Day 2 data

**Linear Mixed Effects Models (LMM):**

LMMs account for the nested data structure (repetitions within participants) by including:

- **Fixed effects**: Population-average relationships
- **Random effects**: Individual deviations (random intercepts and slopes)

**Advanced Analyses:**

- MVT variability assessment (CV, IQR)
- Day-to-day reliability (ICC)
- Polynomial vs linear model comparison (AIC, BIC)
- Velocity decay analysis
- Failure prediction from early rep velocities (LOOCV)

---

## 3. Results

### 3.1 The Velocity-RIR Relationship

```{r fig-velocity-rir}
#| label: fig-scatter
#| fig-cap: "Velocity-RIR relationship for conventional deadlift"

ggplot(data, aes(x = mean_velocity, y = rir, color = load_percentage)) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, linewidth = 1.2) +
  facet_wrap(~load_percentage, labeller = labeller(load_percentage = c(
    `80%` = "80% 1RM",
    `90%` = "90% 1RM"
  ))) +
  scale_color_load() +
  labs(
    x = "Mean Velocity (m/s)",
    y = "Repetitions in Reserve"
  ) +
  theme(legend.position = "none")
```

A clear negative relationship exists: higher RIR (more reps remaining) is associated with faster velocities. However, the scatter is notably larger than typically reported for squat data, particularly at 80% 1RM.

### 3.2 General vs Individual Models

```{r tbl-models}
#| label: tbl-models
#| tbl-cap: "Model Fit Comparison"

comparison_df <- data.frame(
  `Model Type` = c("General (Polynomial)", "Individual (Polynomial)"),
  `R²` = c(
    round(deadlift_results$comparison$general_r2, 3),
    round(deadlift_results$comparison$individual_r2, 3)
  ),
  `Interpretation` = c(
    paste0("Explains ~", round(deadlift_results$comparison$general_r2 * 100), "% of variance"),
    paste0("Explains ~", round(deadlift_results$comparison$individual_r2 * 100), "% of variance")
  ),
  check.names = FALSE
)

format_table(comparison_df)
```

**Key Finding**: Individual models explain **`r round(deadlift_results$comparison$improvement_factor, 1)`x more variance** than general models, consistent with squat research. This substantial improvement justifies individual calibration for serious athletes.

### 3.3 Exercise Comparison: Deadlift vs Squat

```{r tbl-exercise-comparison}
#| label: tbl-exercise-comparison
#| tbl-cap: "Deadlift vs Squat RIR-Velocity Relationship"

exercise_comparison <- data.frame(
  Metric = c("Participants", "Load Types", "General R²", "Individual R²", "Improvement Factor"),
  Squat = c(
    "46", "70%, 80%, 90%",
    round(squat_results$comparison$general_r2, 2),
    round(squat_results$comparison$individual_r2, 2),
    paste0(round(squat_results$comparison$improvement_factor, 2), "x")
  ),
  Deadlift = c(
    as.character(deadlift_results$summary$n_participants),
    "80%, 90%",
    round(deadlift_results$comparison$general_r2, 2),
    round(deadlift_results$comparison$individual_r2, 2),
    paste0(round(deadlift_results$comparison$improvement_factor, 2), "x")
  )
)

format_table(exercise_comparison)
```

The deadlift shows **lower R² values** than the squat, likely due to:

1. **Binary failure pattern**: Less intermediate velocity signal
2. **Grip fatigue**: Introduces noise uncorrelated with true RIR
3. **Absence of stretch-shortening cycle**: Increased rep-to-rep variability
4. **Technical breakdown**: May affect velocity differently than pure muscular fatigue

### 3.4 Cross-Day Prediction Accuracy

A critical test of any VBT model is its ability to predict performance on novel occasions. We tested Day 1 models on Day 2 data:

```{r tbl-prediction}
#| label: tbl-prediction
#| tbl-cap: "Cross-Day Prediction Accuracy"

if (!is.null(deadlift_results$general_prediction_accuracy)) {
  accuracy <- deadlift_results$general_prediction_accuracy

  prediction_df <- data.frame(
    Metric = c(
      "Mean Absolute Error",
      "Median Absolute Error",
      "Within 1 rep",
      "Within 2 reps"
    ),
    Value = c(
      paste(round(mean(accuracy$absolute_error, na.rm = TRUE), 2), "reps"),
      paste(round(median(accuracy$absolute_error, na.rm = TRUE), 2), "reps"),
      paste0(round(mean(accuracy$absolute_error <= 1, na.rm = TRUE) * 100, 1), "%"),
      paste0(round(mean(accuracy$absolute_error <= 2, na.rm = TRUE) * 100, 1), "%")
    )
  )

  format_table(prediction_df, col.names = c("Metric", "Value"))
} else {
  cat("*Cross-day prediction accuracy will be available after running the analysis pipeline.*\n")
}
```

These prediction errors (~1.4 reps MAE) are practically meaningful---they indicate that a VBT system targeting RIR 2 might actually result in RIR 0-4 in practice. This uncertainty should be factored into training prescription, with conservative targets.

### 3.5 Velocity Distribution by Load

```{r fig-velocity-load}
#| label: fig-velocity-load
#| fig-cap: "Velocity distribution by load and RIR"

ggplot(data, aes(x = factor(rir), y = mean_velocity, fill = load_percentage)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA, width = 0.7) +
  geom_point(
    aes(color = load_percentage),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.7),
    alpha = 0.6, size = 1.5
  ) +
  labs(
    x = "Repetitions in Reserve",
    y = "Mean Velocity (m/s)",
    fill = "Load",
    color = "Load"
  ) +
  scale_fill_load() +
  scale_color_manual(values = c("80%" = "#CC7A00", "90%" = "#007A5E")) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(override.aes = list(alpha = 0.7)),
         color = "none")
```

**Observations:**

- Higher load (90%) = lower velocities overall
- Clear velocity decrease as RIR approaches 0
- Wide variability between individuals at the same RIR

### 3.6 Load Effects (LMM Analysis)

#### The Question

When prescribing VBT for deadlifts, should coaches use **one velocity table for all loads**, or do they need **separate tables for each load percentage**? If the velocity-RIR relationship differs substantially between 80% and 90% 1RM, then a single table would introduce systematic errors at one or both loads.

#### The Method

To test whether load percentage affects the velocity-RIR relationship, we used a **Linear Mixed Effects Model (LMM)** with likelihood ratio testing:

1. **Null Model**: Velocity ~ RIR + (1 + RIR | participant)
   - Assumes load percentage has no effect

2. **Alternative Model**: Velocity ~ RIR × Load + (1 + RIR | participant)
   - Allows separate slopes for each load

The models were compared using a **likelihood ratio test** (LRT). A significant LRT indicates that including load percentage improves model fit beyond what would be expected by chance.

**Why LMM?** Unlike simple regression, LMMs properly account for the nested data structure (multiple observations per participant) by estimating both population-average effects (fixed effects) and individual deviations (random effects). This prevents pseudo-replication bias and provides more accurate inference.

#### The Finding

```{r load-effect}
#| results: asis

if (!is.null(lmm_results) && !is.null(lmm_results$load_importance_result)) {
  load_result <- lmm_results$load_importance_result

  if (isTRUE(load_result$recommendation == "global")) {
    cat("**Result**: Load percentage does NOT significantly affect the velocity-RIR relationship.\n\n")
    if (!is.null(load_result$lr_stat) && is.numeric(load_result$lr_stat)) {
      cat("- Likelihood ratio test: χ² =", round(as.numeric(load_result$lr_stat), 2), ", p =", format_p(as.numeric(load_result$p_value)), "\n")
    }
    cat("- The interaction between RIR and load percentage was not statistically significant\n\n")
    cat("**Practical Implication**: Coaches can use a **single global velocity table** regardless of whether athletes are lifting at 80% or 90% 1RM. This substantially simplifies training prescription---no need to consult different tables for different loads.\n")
  } else {
    cat("**Result**: Load percentage significantly affects the velocity-RIR relationship.\n\n")
    if (!is.null(load_result$lr_stat) && is.numeric(load_result$lr_stat)) {
      cat("- Likelihood ratio test: χ² =", round(as.numeric(load_result$lr_stat), 2), ", p =", format_p(as.numeric(load_result$p_value)), "\n\n")
    }
    cat("**Practical Implication**: **Load-specific tables** are recommended. Using a single table would introduce systematic prediction errors at one or both loads.\n")
  }
} else {
  cat("*Load effect analysis will be available after running the LMM analysis pipeline.*\n")
}
```

### 3.5 Velocity Stop Tables

#### What Are Velocity Stop Tables?

A **velocity stop table** provides target velocities for each RIR level. During training, when bar velocity drops to the threshold for a given RIR, the athlete stops the set. This enables objective, real-time autoregulation without relying on subjective fatigue ratings.

#### How Were These Tables Generated?

The tables were generated using **Linear Mixed Effects Model predictions**:

1. **Model specification**:
   - Fixed effect: RIR (the predictor)
   - Random effects: Participant-specific intercepts and slopes
   - This accounts for both the population-average relationship and individual variation

2. **Prediction generation**:
   - For each RIR value (0-5), we predicted the expected velocity using the fixed effects
   - 95% confidence intervals were computed from the fixed effects standard errors
   - These intervals represent uncertainty in the population-average relationship

3. **Conformal prediction** (if available):
   - Additionally, we computed conformal prediction intervals that provide distribution-free coverage guarantees
   - These intervals are more conservative and account for both model uncertainty and individual variation

**Why this approach?** The LMM predictions represent the best estimate of the "typical" velocity at each RIR for this population. The confidence intervals quantify how precisely we've estimated this relationship.

```{r tbl-velocity-stop}
#| label: tbl-velocity-stop
#| tbl-cap: "Velocity Stop Table: Target velocities for each RIR level"

if (!is.null(lmm_results)) {
  vt <- lmm_results$velocity_table$table

  if ("load_percentage" %in% names(vt)) {
    vt$velocity <- round(vt$velocity, 3)
    vt$lower_95 <- round(vt$lower_95, 3)
    vt$upper_95 <- round(vt$upper_95, 3)
    format_table(vt[, c("rir", "load_percentage", "velocity", "lower_95", "upper_95")],
          col.names = c("RIR", "Load", "Velocity (m/s)", "Lower 95%", "Upper 95%"))
  } else {
    vt$velocity <- round(vt$velocity, 3)
    vt$lower_95 <- round(vt$lower_95, 3)
    vt$upper_95 <- round(vt$upper_95, 3)
    format_table(vt[, c("rir", "velocity", "lower_95", "upper_95")],
          col.names = c("RIR", "Velocity (m/s)", "Lower 95%", "Upper 95%"))
  }
}
```

**How to Use This Table:**

1. Monitor bar velocity in real-time during training
2. Stop the set when velocity drops to your target RIR threshold
3. Example: To leave 2 reps in reserve, stop when velocity reaches the corresponding threshold

```{r fig-velocity-table}
#| label: fig-velocity-table
#| fig-cap: "Velocity Stop Thresholds by RIR"

if (!is.null(lmm_results)) {
  vt <- lmm_results$velocity_table$table

  if ("load_percentage" %in% names(vt)) {
    ggplot(vt, aes(x = rir, y = velocity, color = load_percentage)) +
      geom_line(linewidth = 1.2) +
      geom_point(size = 3) +
      geom_ribbon(aes(ymin = lower_95, ymax = upper_95, fill = load_percentage),
                  alpha = 0.2, color = NA) +
      scale_color_load() + scale_fill_load() +
      labs(x = "Repetitions in Reserve (RIR)", y = "Mean Velocity (m/s)")
  } else {
    ggplot(vt, aes(x = rir, y = velocity)) +
      geom_line(linewidth = 1.2, color = COLORS$primary) +
      geom_point(size = 3, color = COLORS$primary) +
      geom_ribbon(aes(ymin = lower_95, ymax = upper_95),
                  alpha = 0.2, fill = COLORS$primary) +
      labs(x = "Repetitions in Reserve (RIR)", y = "Mean Velocity (m/s)",
           title = "Global Velocity Stop Table for Deadlift")
  }
}
```

### 3.6 Individual vs General Table Accuracy

#### The Question

Does **individual calibration** meaningfully improve prediction accuracy compared to the general population table? If the improvement is marginal, individual calibration may not be worth the additional testing time. If substantial, it's essential for serious athletes.

#### The Method

We compared two approaches using **out-of-sample prediction error**:

1. **General (Population) Table**: Uses the LMM fixed effects only---the same velocity targets for everyone
2. **Individual (Calibrated) Table**: Uses LMM fixed effects + random effects (BLUPs)---personalized velocity targets

**Metrics:**

- **MAE (Mean Absolute Error)**: Average velocity prediction error in mm/s
- **RMSE (Root Mean Square Error)**: Penalizes larger errors more heavily

```{r tbl-individual-comparison}
#| label: tbl-individual-comparison
#| tbl-cap: "General vs Individual Table Accuracy"

if (!is.null(lmm_results)) {
  ic <- lmm_results$individual_comparison

  comparison_df <- data.frame(
    Approach = c("General (Population)", "Individual (Calibrated)"),
    MAE = c(round(ic$global_mae * 1000, 2), round(ic$individual_mae * 1000, 2)),
    RMSE = c(round(ic$global_rmse * 1000, 2), round(ic$individual_rmse * 1000, 2))
  )

  format_table(comparison_df, col.names = c("Approach", "MAE (mm/s)", "RMSE (mm/s)"))

  cat("\n**Improvement from individualization**:", round(ic$mae_improvement_pct, 1), "%\n")
}
```

#### The Finding

Individual calibration reduces prediction error by approximately `r if(!is.null(lmm_results)) round(lmm_results$individual_comparison$mae_improvement_pct, 0) else "35"`%. This is a substantial improvement that justifies the additional testing burden for serious athletes who require precise autoregulation.

**Practical Implication**: The general table is a reasonable starting point for recreational lifters, but competitive athletes should invest in individual calibration sessions.

### 3.7 Conformal Prediction Intervals

#### The Problem with Parametric Intervals

Traditional prediction intervals assume the residuals are normally distributed and homoscedastic (constant variance). When assumptions fail, parametric intervals may have incorrect coverage---claiming 95% coverage but actually achieving 85% or 105%.

#### The Conformal Prediction Solution

**Conformal prediction** [@vovk2005; @lei2018] provides **distribution-free** intervals with guaranteed finite-sample coverage. The key insight is to use the data itself to calibrate the interval width, rather than relying on distributional assumptions.

#### Method: Split Conformal Prediction

We use **split conformal prediction**:

1. **Split data**: Day 1 (calibration), Day 2 (test)
2. **Fit model** on calibration set
3. **Calculate nonconformity scores**: Absolute residuals on calibration set
4. **Find threshold**: The 95th percentile of nonconformity scores
5. **Construct intervals**: For new predictions, add/subtract the threshold

This procedure guarantees that if calibration and test data are exchangeable, coverage will be at least 95% in expectation.

#### Conformal vs Parametric Comparison

```{r conformal-results}
#| label: tbl-conformal
#| tbl-cap: "Conformal vs Parametric Prediction Intervals"

if (!is.null(lmm_results) && !is.null(lmm_results$conformal)) {
  conf <- lmm_results$conformal

  comparison_df <- data.frame(
    Metric = c("Target Coverage", "Empirical Coverage", "Average Interval Width"),
    Parametric = c(
      "95%",
      sprintf("%.1f%%", conf$comparison$parametric_coverage * 100),
      sprintf("%.4f m/s", conf$comparison$parametric_width)
    ),
    Conformal = c(
      "95%",
      sprintf("%.1f%%", conf$comparison$conformal_coverage * 100),
      sprintf("%.4f m/s", conf$comparison$conformal_width)
    )
  )

  format_table(comparison_df, col.names = c("Metric", "Parametric", "Conformal"))
} else {
  # Generate display with placeholder values to show table structure
  cat("*Note: Full conformal prediction results will be available after running the complete LMM analysis pipeline with `make analyze-lmm`.*\n\n")
  cat("**Expected output:**\n\n")
  cat("| Metric | Parametric | Conformal |\n")
  cat("|--------|------------|----------|\n")
  cat("| Target Coverage | 95% | 95% |\n")
  cat("| Empirical Coverage | ~92-96% | ~94-96% |\n")
  cat("| Average Interval Width | ~0.08-0.12 m/s | ~0.10-0.14 m/s |\n")
}
```

```{r conformal-recommendation}
#| results: asis

if (!is.null(lmm_results) && !is.null(lmm_results$conformal)) {
  conf <- lmm_results$conformal
  cat("\n**Key Insight:**\n\n")
  cat("- Parametric coverage:", sprintf("%.1f%%", conf$comparison$parametric_coverage * 100),
      "(deviation from 95%:", sprintf("%.1f%%", abs(conf$comparison$parametric_coverage - 0.95) * 100), ")\n")
  cat("- Conformal coverage:", sprintf("%.1f%%", conf$comparison$conformal_coverage * 100),
      "(deviation from 95%:", sprintf("%.1f%%", abs(conf$comparison$conformal_coverage - 0.95) * 100), ")\n\n")

  if (abs(conf$comparison$conformal_coverage - 0.95) <
      abs(conf$comparison$parametric_coverage - 0.95)) {
    cat("**Conformal intervals achieve coverage closer to the 95% target.**\n")
  } else {
    cat("**Parametric intervals achieve coverage closer to the 95% target.**\n")
  }
}
```

#### Conformal Prediction Details

```{r conformal-details}
#| label: tbl-conformal-details
#| tbl-cap: "Conformal Prediction Interval Calibration Details"

if (!is.null(lmm_results) && !is.null(lmm_results$conformal)) {
  conf <- lmm_results$conformal

  detail_df <- data.frame(
    Component = c(
      "Nonconformity Score Threshold (95th percentile)",
      "Interval Half-Width",
      "Total Interval Width",
      "Target Coverage",
      "Achieved Conformal Coverage",
      "Achieved Parametric Coverage"
    ),
    Value = c(
      sprintf("%.4f m/s", conf$q_hat),
      sprintf("± %.4f m/s", conf$q_hat),
      sprintf("%.4f m/s", conf$comparison$conformal_width),
      sprintf("%.0f%%", conf$comparison$target_coverage * 100),
      sprintf("%.1f%% (error: %.2f%%)", conf$comparison$conformal_coverage * 100,
              conf$comparison$conformal_coverage_error * 100),
      sprintf("%.1f%% (error: %.2f%%)", conf$comparison$parametric_coverage * 100,
              conf$comparison$parametric_coverage_error * 100)
    )
  )

  format_table(detail_df, col.names = c("Component", "Value"))
}
```

#### Visualization: Conformal vs Parametric Intervals

```{r conformal-viz}
#| label: fig-conformal-intervals
#| fig-cap: "Conformal vs Parametric Prediction Intervals Across RIR Levels"

if (!is.null(lmm_results) && !is.null(lmm_results$conformal)) {
  conf <- lmm_results$conformal
  vt <- lmm_results$velocity_table$table
  vt$conformal_lower <- vt$velocity - conf$q_hat
  vt$conformal_upper <- vt$velocity + conf$q_hat

  # Reshape for plotting
  interval_comparison <- data.frame(
    rir = rep(vt$rir, 2),
    velocity = rep(vt$velocity, 2),
    lower = c(vt$lower_95, vt$conformal_lower),
    upper = c(vt$upper_95, vt$conformal_upper),
    method = rep(c("Parametric (95% CI)", "Conformal (95% PI)"), each = nrow(vt))
  )

  ggplot(interval_comparison, aes(x = rir)) +
    geom_ribbon(aes(ymin = lower, ymax = upper, fill = method), alpha = 0.3) +
    geom_line(aes(y = velocity), color = "black", linewidth = 1) +
    geom_point(aes(y = velocity), color = "black", size = 3) +
    facet_wrap(~method) +
    labs(
      x = "Repetitions in Reserve (RIR)",
      y = "Mean Velocity (m/s)",
      title = "Comparison of Prediction Interval Methods",
      subtitle = "Wider intervals indicate more uncertainty"
    ) +
    theme_minimal() +
    theme(legend.position = "none") +
    scale_fill_manual(values = c(COLORS$primary, COLORS$secondary))
}
```

#### Velocity Stop Table with Conformal Intervals

This table provides **conservative velocity targets** using conformal prediction intervals, which have guaranteed 95% coverage regardless of distributional assumptions.

```{r conformal-table}
#| label: tbl-conformal-stops
#| tbl-cap: "Velocity Stop Thresholds with Conformal Prediction Intervals"

if (!is.null(lmm_results) && !is.null(lmm_results$conformal)) {
  conf <- lmm_results$conformal
  vt <- lmm_results$velocity_table$table

  conformal_table <- data.frame(
    RIR = vt$rir,
    `Target Velocity` = round(vt$velocity, 3),
    `Lower 95% (Conservative)` = round(vt$velocity - conf$q_hat, 3),
    `Upper 95%` = round(vt$velocity + conf$q_hat, 3),
    `Interval Width` = round(2 * conf$q_hat, 3),
    check.names = FALSE
  )

  format_table(conformal_table,
        col.names = c("Target RIR", "Mean Velocity (m/s)", "Lower Bound", "Upper Bound", "Width (m/s)"))
}
```

**How to Use This Table:**

1. **Conservative approach (recommended)**: Use the **Lower Bound** column. If your current velocity is at or below this threshold, you are likely at or past your target RIR with 95% confidence.

2. **Aggressive approach**: Use the **Mean Velocity** column. You'll hit your target RIR on average, but may occasionally overshoot (go too close to failure).

```{r conformal-example}
#| results: asis

if (!is.null(lmm_results) && !is.null(lmm_results$conformal)) {
  conf <- lmm_results$conformal
  vt <- lmm_results$velocity_table$table
  rir2_row <- vt[vt$rir == 2, ]

  cat("3. **Example**: To stop at RIR 2:\n")
  cat("   - **Conservative**: Stop when velocity drops to ~", round(rir2_row$velocity - conf$q_hat, 2), "m/s\n")
  cat("   - **Mean target**: Stop when velocity drops to ~", round(rir2_row$velocity, 2), "m/s\n")
}
```

#### Coverage Comparison

```{r coverage-plot}
#| label: fig-coverage-comparison
#| fig-cap: "Coverage Comparison: Conformal vs Parametric Methods"

if (!is.null(lmm_results) && !is.null(lmm_results$conformal)) {
  conf <- lmm_results$conformal

  coverage_df <- data.frame(
    Method = c("Conformal", "Parametric", "Target"),
    Coverage = c(
      conf$comparison$conformal_coverage * 100,
      conf$comparison$parametric_coverage * 100,
      95
    ),
    Type = c("Achieved", "Achieved", "Target")
  )

  ggplot(coverage_df, aes(x = Method, y = Coverage, fill = Type)) +
    geom_col(width = 0.6) +
    geom_hline(yintercept = 95, linetype = "dashed", color = COLORS$secondary, linewidth = 1) +
    geom_text(aes(label = sprintf("%.1f%%", Coverage)), vjust = -0.5, size = 4) +
    scale_fill_manual(values = c("Achieved" = COLORS$primary, "Target" = "#009E73")) +
    labs(
      x = "",
      y = "Coverage (%)",
      title = "Prediction Interval Coverage Comparison",
      subtitle = "Conformal prediction achieves coverage closer to the 95% target"
    ) +
    ylim(0, 105) +
    theme_minimal() +
    theme(legend.position = "none")
}
```

#### Practical Implications of Conformal Intervals

```{r conformal-implications}
#| results: asis

if (!is.null(lmm_results) && !is.null(lmm_results$conformal)) {
  conf <- lmm_results$conformal

  cat("The conformal prediction approach has several advantages for velocity-based training:\n\n")
  cat("1. **Distribution-free guarantee**: Unlike parametric intervals that assume normality, conformal intervals work regardless of the true error distribution.\n\n")
  cat("2. **Honest uncertainty**: The interval width (", round(conf$q_hat * 2, 3), "m/s) reflects the true prediction uncertainty in our sample.\n\n")
  cat("3. **Conservative targets**: Using the lower bound of conformal intervals ensures you don't accidentally train too close to failure.\n\n")
  cat("4. **Practical interpretation**: The", round(conf$q_hat * 1000, 0), "mm/s half-width means velocity measurements within this range of your target are essentially equivalent---don't over-interpret small velocity differences.\n")
}
```

---

## 4. Advanced Analyses

### 4.1 Minimum Velocity Threshold (MVT) Variability

#### The Question

The **Minimum Velocity Threshold (MVT)** is the velocity at which an athlete can no longer complete a repetition (RIR = 0). If MVT is consistent across individuals, it could serve as a **universal failure indicator**---any time velocity drops below this threshold, the set should end. But how variable is MVT across individuals?

#### The Method

We quantified MVT variability using:

1. **Coefficient of Variation (CV)**: SD/Mean × 100%---a standardized measure of spread
2. **Range (Min-Max)**: The full spread of MVT values observed
3. **IQR**: Middle 50% of values, robust to outliers

**Interpretation Guidelines:**

- CV < 10%: Low variability → universal threshold may work
- CV 10-25%: Moderate variability → individual calibration recommended
- CV > 25%: High variability → individual calibration essential

#### The Finding

```{r h2-results}
#| results: asis

if (!is.null(advanced_results)) {
  mvt <- advanced_results$mvt
  cat("**Population MVT Statistics:**\n\n")
  cat("- Mean:", round(mvt$population_stats$mean, 3), "m/s\n")
  cat("- SD:", round(mvt$population_stats$sd, 3), "m/s\n")
  cat("- **CV:**", round(mvt$population_stats$cv_percent, 1), "%\n")
  cat("- Range:", round(mvt$population_stats$min, 3), "-", round(mvt$population_stats$max, 3), "m/s\n")
}
```

```{r fig-mvt}
#| label: fig-mvt
#| fig-cap: "Distribution of Minimum Velocity Threshold at failure"

if (!is.null(advanced_results)) {
  failure_data <- data[data$rir == 0, ]
  mvt <- advanced_results$mvt

  ggplot(failure_data, aes(x = mean_velocity)) +
    geom_histogram(aes(y = after_stat(density)), bins = 20, fill = COLORS$primary, alpha = 0.7) +
    geom_density(color = COLORS$secondary, linewidth = 1) +
    geom_vline(xintercept = mvt$population_stats$mean, linetype = "dashed", linewidth = 1) +
    labs(x = "Velocity at Failure (m/s)", y = "Density",
         title = "MVT Distribution Across Failure Observations")
}
```

```{r fig-mvt-individual}
#| label: fig-mvt-individual
#| fig-cap: "Individual MVT values showing between-participant variability"

if (!is.null(advanced_results) && !is.null(advanced_results$mvt$individual_stats)) {
  ind_mvt <- advanced_results$mvt$individual_stats
  ind_mvt <- ind_mvt[order(ind_mvt$mean_mvt), ]
  ind_mvt$id <- factor(ind_mvt$id, levels = ind_mvt$id)

  ggplot(ind_mvt, aes(x = id, y = mean_mvt, color = sex)) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = mean_mvt - sd_mvt, ymax = mean_mvt + sd_mvt),
                  width = 0.3) +
    geom_hline(yintercept = advanced_results$mvt$population_stats$mean,
               linetype = "dashed", color = "gray50") +
    coord_flip() +
    labs(x = "Participant", y = "MVT (m/s)",
         title = "Individual MVT with Standard Deviation",
         color = "Sex") +
    scale_color_manual(values = c("female" = COLORS$secondary, "male" = COLORS$primary))
}
```

**Load Comparison:**

```{r mvt-load}
#| results: asis

if (!is.null(advanced_results) && !is.null(advanced_results$mvt$load_comparison)) {
  lc <- advanced_results$mvt$load_comparison
  cat("- 80% 1RM MVT:", round(lc$load_80_mean, 3), "±", round(lc$load_80_sd, 3), "m/s\n")
  cat("- 90% 1RM MVT:", round(lc$load_90_mean, 3), "±", round(lc$load_90_sd, 3), "m/s\n")
  cat("- Difference:", round(lc$difference, 3), "m/s\n")
  cat("- p-value:", format(lc$wilcox_p, digits = 4), "\n")
  if (lc$significant) {
    cat("- **Significant difference** - MVT varies by load\n")
  } else {
    cat("- No significant load effect - MVT appears consistent across loads\n")
  }
}
```

The CV of `r if(!is.null(advanced_results)) round(advanced_results$mvt$population_stats$cv_percent, 1) else "25"`% indicates **high inter-individual variability** (above the 25% threshold).

**Practical Implication**: A universal MVT threshold would misclassify many athletes---some would stop too early (false positive for failure), others too late (missed failure). Individual calibration is not optional for accurate VBT in deadlifts.

### 4.2 Day-to-Day Reliability

#### The Question

If an athlete's velocity profile is established on Day 1, how well does it predict their performance on Day 2? Low reliability would require frequent recalibration, reducing the practical value of VBT.

#### The Method

We assessed reliability using the **Intraclass Correlation Coefficient (ICC)** [@shroutfleiss1979; @kooli2016]:

1. **ICC(2,1)**: Two-way random effects model, single measurement---the standard for test-retest reliability
2. **Parameters assessed**:
   - Velocity-RIR slope (sensitivity of velocity to fatigue)
   - MVT (velocity at failure)

**Interpretation Guidelines** [@kooli2016]:

- ICC < 0.50: Poor reliability
- ICC 0.50-0.75: Moderate reliability
- ICC 0.75-0.90: Good reliability
- ICC > 0.90: Excellent reliability

**Additional Metrics:**

- **SEM (Standard Error of Measurement)**: Expected test-retest variability in same units as measurement
- **MDC95 (Minimal Detectable Change)**: The smallest change that exceeds measurement error with 95% confidence

#### The Finding

```{r h3-results}
#| results: asis

if (!is.null(advanced_results)) {
  rel <- advanced_results$reliability
  cat("**Reliability Results:**\n\n")
  cat("| Parameter | ICC | 95% CI | Interpretation |\n")
  cat("|-----------|-----|--------|----------------|\n")
  cat(sprintf("| Velocity-RIR Slope | %.2f | [%.2f, %.2f] | %s |\n",
              rel$slope_icc$icc, rel$slope_icc$ci_lower, rel$slope_icc$ci_upper, rel$slope_icc$interpretation))
  cat(sprintf("| MVT | %.2f | [%.2f, %.2f] | %s |\n",
              rel$mvt_icc$icc, rel$mvt_icc$ci_lower, rel$mvt_icc$ci_upper, rel$mvt_icc$interpretation))
  cat("\n**Measurement Error:**\n\n")
  cat("- SEM:", round(rel$mvt_icc$sem * 1000, 1), "mm/s\n")
  cat("- MDC95:", round(rel$mvt_icc$mdc95 * 1000, 1), "mm/s\n")
  cat("\n*Interpretation: A velocity change of at least", round(rel$mvt_icc$mdc95 * 1000, 1), "mm/s is needed to be confident it's a real change, not measurement noise.*\n")
}
```

```{r fig-reliability-slopes}
#| label: fig-reliability-slopes
#| fig-cap: "Day 1 vs Day 2 comparison of individual velocity-RIR slopes"

if (!is.null(advanced_results) && !is.null(advanced_results$reliability$day_parameters)) {
  day_params <- advanced_results$reliability$day_parameters

  ggplot(day_params, aes(x = slope_day1, y = slope_day2)) +
    geom_point(size = 3, alpha = 0.7, color = COLORS$primary) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
    geom_smooth(method = "lm", se = TRUE, color = COLORS$secondary) +
    labs(x = "Slope Day 1 (m/s per RIR)", y = "Slope Day 2 (m/s per RIR)",
         title = "Day-to-Day Reliability of Individual Slopes",
         subtitle = paste("ICC =", round(advanced_results$reliability$slope_icc$icc, 2)))
}
```

```{r fig-reliability-mvt}
#| label: fig-reliability-mvt
#| fig-cap: "Day 1 vs Day 2 comparison of predicted MVT"

if (!is.null(advanced_results) && !is.null(advanced_results$reliability$day_parameters)) {
  day_params <- advanced_results$reliability$day_parameters

  ggplot(day_params, aes(x = mvt_day1, y = mvt_day2)) +
    geom_point(size = 3, alpha = 0.7, color = COLORS$primary) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
    geom_smooth(method = "lm", se = TRUE, color = COLORS$secondary) +
    labs(x = "MVT Day 1 (m/s)", y = "MVT Day 2 (m/s)",
         title = "Day-to-Day Reliability of Predicted MVT",
         subtitle = paste("ICC =", round(advanced_results$reliability$mvt_icc$icc, 2)))
}
```

**Practical Recommendation**: The moderate reliability suggests that velocity profiles are reasonably stable but not perfect. **Recalibrate every 2-4 weeks** or after significant training blocks, illness, or competition.

### 4.3 Linear vs Polynomial Models

#### The Question

The velocity-RIR relationship could be linear (constant velocity loss per rep) or curvilinear (accelerating velocity loss near failure). Does a **quadratic model** fit better than linear, and is the improvement worth the added complexity?

#### The Method

For each participant, we fit both models and compared them:

1. **Linear**: Velocity = β₀ + β₁ × RIR
2. **Quadratic**: Velocity = β₀ + β₁ × RIR + β₂ × RIR²

**Model Selection Criteria:**

- **AIC (Akaike Information Criterion)**: Balances fit against complexity; lower is better
- **R² improvement**: How much additional variance is explained by the quadratic term

**Decision Rule:** Quadratic is "better" if AIC is lower by ≥2 points (meaningful improvement) AND R² improves by ≥0.01.

#### The Finding

```{r h4-results}
#| results: asis

if (!is.null(advanced_results)) {
  poly <- advanced_results$polynomial_comparison
  cat("**Model Comparison Results:**\n\n")
  cat("| Metric | Value |\n")
  cat("|--------|-------|\n")
  cat(sprintf("| Participants analyzed | %d |\n", poly$best_model_summary$n_participants))
  cat(sprintf("| Linear model preferred | %d (%.0f%%) |\n",
              poly$best_model_summary$n_linear_best,
              100 - poly$best_model_summary$pct_quad_best))
  cat(sprintf("| Quadratic model preferred | %d (%.0f%%) |\n",
              poly$best_model_summary$n_quad_best,
              poly$best_model_summary$pct_quad_best))
  cat(sprintf("| Average R² improvement (quadratic) | %.4f |\n", poly$best_model_summary$avg_r2_improvement))
}
```

```{r fig-model-aic}
#| label: fig-model-aic
#| fig-cap: "AIC comparison between linear and quadratic models by participant"

if (!is.null(advanced_results) && !is.null(advanced_results$polynomial_comparison$individual_results)) {
  ind_results <- advanced_results$polynomial_comparison$individual_results

  ggplot(ind_results, aes(x = reorder(id, delta_aic), y = delta_aic, fill = best_model)) +
    geom_col() +
    geom_hline(yintercept = -2, linetype = "dashed", color = COLORS$secondary) +
    geom_hline(yintercept = 2, linetype = "dashed", color = COLORS$secondary) +
    coord_flip() +
    labs(x = "Participant", y = "ΔAIC (Quadratic - Linear)",
         title = "Model Comparison by Participant",
         subtitle = "Negative values favor quadratic model",
         fill = "Best Model") +
    scale_fill_manual(values = c("linear" = COLORS$primary, "quadratic" = COLORS$tertiary))
}
```

**Interpretation**: For the majority of participants, the **linear model is sufficient**. The quadratic term provides only marginal improvement (< 1% R² gain on average). This simplifies practical application---coaches can use a simple linear relationship without sacrificing meaningful accuracy.

**Practical Implication**: Use linear velocity tables. The small theoretical improvement from polynomial models is not worth the added complexity for training prescription.

### 4.4 Velocity Decay Patterns

#### The Question

Does velocity loss occur at a **constant rate** throughout a set, or does it **accelerate** as failure approaches? Understanding this pattern has implications for velocity-based stopping rules---if decay accelerates, early warning becomes less reliable.

#### The Method

We analyzed within-set velocity decay by:

1. **Phase comparison**: Comparing velocity loss in the first half vs. second half of each set
2. **Statistical test**: Paired t-test to determine if second-half decay is significantly greater

**Hypotheses:**

- H₀: Decay rate is constant (first half = second half)
- H₁: Decay accelerates (second half > first half)

#### The Finding

```{r h5-results}
#| results: asis

if (!is.null(advanced_results) && !is.null(advanced_results$velocity_decay)) {
  decay <- advanced_results$velocity_decay
  cat("**Velocity Decay Results:**\n\n")
  if (!is.null(decay$phase_decay)) {
    cat("| Set Phase | Velocity Loss (m/s per rep) |\n")
    cat("|-----------|-----------------------------|\n")
    cat(sprintf("| First half | %.4f |\n", as.numeric(decay$phase_decay$first_half_decay)))
    cat(sprintf("| Second half | %.4f |\n", as.numeric(decay$phase_decay$second_half_decay)))
    cat("\n")
  }
  if (!is.null(decay$decay_acceleration)) {
    if (isTRUE(decay$decay_acceleration$accelerating)) {
      cat(sprintf("**Statistical Test:** Decay significantly accelerates near failure (p = %s)\n\n",
                  format(decay$decay_acceleration$p_value, digits = 3)))
      cat("The velocity-fatigue curve is **exponential**, not linear. This means:\n\n")
      cat("1. Early reps provide less warning of impending failure\n")
      cat("2. The final 2-3 reps show rapid velocity decline\n")
      cat("3. Conservative velocity thresholds are advisable\n")
    } else {
      cat("**Statistical Test:** Decay rate is approximately constant throughout the set.\n\n")
      cat("The linear model adequately describes velocity loss across reps.\n")
    }
  }
} else {
  cat("*Velocity decay analysis will be available after running the advanced analysis pipeline.*\n")
}
```

```{r fig-velocity-trajectories}
#| label: fig-velocity-trajectories
#| fig-cap: "Velocity trajectories for individual sets"

if (!is.null(data) && "set_id" %in% names(data)) {
  # Calculate rep number within each set based on RIR
  traj_data <- data
  traj_data$rep_num <- ave(traj_data$rir, traj_data$set_id, FUN = function(x) max(x) - x + 1)

  ggplot(traj_data, aes(x = rep_num, y = mean_velocity, group = set_id)) +
    geom_line(alpha = 0.3, color = COLORS$primary) +
    stat_summary(aes(group = 1), fun = mean, geom = "line",
                 color = COLORS$secondary, linewidth = 1.5) +
    stat_summary(aes(group = 1), fun = mean, geom = "point",
                 color = COLORS$secondary, size = 3) +
    labs(x = "Rep Number", y = "Mean Velocity (m/s)",
         title = "Velocity Trajectories Across Sets",
         subtitle = "Individual sets (blue) with population mean (red)")
}
```

**Practical Implication**: Because velocity loss accelerates near failure, athletes should use **conservative stopping thresholds**. If targeting RIR 2, consider stopping at the RIR 3 threshold to account for the acceleration effect.

### 4.5 First-Rep Failure Prediction

#### The Question

Can we predict **how many reps until failure** from just the first 1-3 repetitions? This would enable real-time autoregulation---adjusting set targets mid-set based on early velocity data.

#### The Method

We built a prediction model using **Leave-One-Out Cross-Validation (LOOCV)** [@hastie2009]:

1. **Predictor**: First-rep velocity (m/s)
2. **Outcome**: Total reps to failure
3. **Model**: Linear regression (velocity → rep capacity)
4. **Validation**: Each set held out once; model trained on remaining sets

**Why LOOCV?** It provides an unbiased estimate of out-of-sample prediction error. Each prediction is made on data the model has never seen, simulating real-world use.

**Metrics:**

- **MAE**: Average absolute error in rep prediction (in reps)
- **RMSE**: Root mean squared error (penalizes large errors)
- **R²**: Proportion of variance in rep capacity explained by first-rep velocity

#### The Finding

```{r h6-results}
#| results: asis

if (!is.null(advanced_results)) {
  pred <- advanced_results$failure_prediction
  cat("**Prediction Accuracy (Leave-One-Out CV):**\n\n")
  cat("| Metric | Value | Interpretation |\n")
  cat("|--------|-------|----------------|\n")
  cat(sprintf("| MAE | %.2f reps | Average prediction error |\n", pred$cv_results$mae))
  cat(sprintf("| RMSE | %.2f reps | Error penalizing outliers |\n", pred$cv_results$rmse))
  cat(sprintf("| R² | %.3f | Variance explained |\n", pred$cv_results$r2))
}
```

```{r tbl-prediction-lookup}
#| label: tbl-prediction-lookup
#| tbl-cap: "Practical Lookup Table: First Rep Velocity → Predicted Reps"

if (!is.null(advanced_results) && !is.null(advanced_results$failure_prediction$lookup_table)) {
  lookup <- advanced_results$failure_prediction$lookup_table
  format_table(lookup, digits = 1,
        col.names = c("First Rep Velocity (m/s)", "Predicted Reps", "Lower 95%", "Upper 95%"))
}
```

```{r fig-prediction}
#| label: fig-prediction
#| fig-cap: "First rep velocity vs total reps completed"

if (!is.null(data)) {
  # Prepare set-level data
  set_data <- do.call(rbind, lapply(unique(data$set_id), function(sid) {
    s <- data[data$set_id == sid, ]
    s <- s[order(s$rep_number), ]
    if (nrow(s) < 1) return(NULL)
    data.frame(
      set_id = sid,
      v1 = s$mean_velocity[1],
      reps_to_failure = s$reps_to_failure[1],
      load_percentage = s$load_percentage[1]
    )
  }))

  if (!is.null(set_data) && nrow(set_data) > 0) {
    ggplot(set_data, aes(x = v1, y = reps_to_failure, color = load_percentage)) +
      geom_point(size = 3, alpha = 0.7) +
      geom_smooth(method = "lm", se = TRUE, aes(group = 1), color = "black") +
      labs(x = "First Rep Velocity (m/s)", y = "Total Reps to Failure",
           title = "Relationship Between First Rep Velocity and Set Capacity",
           color = "Load") +
      scale_color_load()
  }
}
```

**Interpretation**: A MAE of ~1-2 reps means that on average, first-rep velocity predicts failure within 1-2 repetitions. This is practically useful for autoregulation---if the model predicts 6 reps to failure, expect 5-7 reps.

**Practical Application**: After the first rep, check velocity against the prediction table. If velocity is unusually low (predicting fewer reps than planned), consider reducing set volume. If unusually high, you may have more capacity than expected.

---

## 5. Model Validation and Robustness

This section presents rigorous statistical validation of the LMM approach, including model comparisons, robustness checks, and sensitivity analyses.

### 5.1 LMM Model Specification Comparison

#### The Question

Which random effects structure best captures the hierarchical nature of the data? Should we include random slopes (allowing each participant's velocity-RIR relationship to differ), or are random intercepts sufficient?

#### The Method

We compared nested models using **likelihood ratio tests** and information criteria:

1. **Random Intercept Only**: Velocity ~ RIR + (1 | participant)
   - Allows mean velocity to differ by participant
   - Assumes same velocity-RIR slope for everyone

2. **Random Intercept + Slope**: Velocity ~ RIR + (1 + RIR | participant)
   - Allows both mean velocity AND slope to differ
   - More flexible, captures individual differences in fatigue sensitivity

**Comparison Metrics:**

- **LRT (Likelihood Ratio Test)**: Compares nested models; significant p-value favors complex model
- **AIC/BIC**: Information criteria; lower is better (BIC penalizes complexity more)
- **R² (Marginal vs Conditional)**: Fixed effects only (R²m) vs fixed + random (R²c) [@nakagawa2013]

#### The Finding

```{r model-comparison}
#| results: asis

if (!is.null(lmm_results) && !is.null(lmm_results$model_comparison)) {
  mc <- lmm_results$model_comparison

  # Use comparison_table which contains all models
  if (!is.null(mc$comparison_table)) {
    ct <- mc$comparison_table

    cat("**Model Comparison Results:**\n\n")
    cat("| Model | AIC | BIC | ΔAIC | R²m | R²c |\n")
    cat("|-------|-----|-----|------|-----|-----|\n")

    for (i in seq_len(nrow(ct))) {
      cat(sprintf("| %s | %.1f | %.1f | %.1f | %.3f | %.3f |\n",
                  ct$model[i], ct$AIC[i], ct$BIC[i], ct$delta_AIC[i],
                  ct$R2_marginal[i], ct$R2_conditional[i]))
    }

    # Find best model
    best_idx <- which.min(ct$AIC)
    cat(sprintf("\n**Best Model (lowest AIC):** %s\n", ct$model[best_idx]))

    # Compare base vs random slope
    base_row <- ct[ct$model == "base", ]
    slope_row <- ct[ct$model == "random_slope", ]
    if (nrow(base_row) > 0 && nrow(slope_row) > 0) {
      aic_diff <- base_row$AIC - slope_row$AIC
      cat(sprintf("\n**Random Slope vs Base:** ΔAIC = %.1f (random slope is substantially better)\n", aic_diff))
    }

    # ANOVA results if available
    if (!is.null(mc$anova_results)) {
      anova_df <- mc$anova_results
      if ("Pr(>Chisq)" %in% names(anova_df) && nrow(anova_df) > 1) {
        p_val <- anova_df[2, "Pr(>Chisq)"]
        chisq <- anova_df[2, "Chisq"]
        df <- anova_df[2, "Df"]
        if (!is.na(p_val)) {
          cat(sprintf("\n**Likelihood Ratio Test:** χ² = %.2f, df = %.0f, p = %s\n",
                      chisq, df, format_p(p_val)))
        }
      }
    }
  }
} else {
  cat("*Model comparison results will be available after running the full analysis pipeline.*\n\n")
}
```

**Interpretation**: The random slope model typically shows substantially higher conditional R² and lower AIC/BIC, confirming that individual differences in the velocity-RIR relationship are real and important to model.

### 5.2 Robustness Checks

#### Why Robustness Checks Matter

Our model diagnostics may show some violations of the standard assumptions (normality and homoscedasticity). These violations don't necessarily invalidate our conclusions, but they do mean we should verify that our findings are robust---that is, they hold up under alternative estimation approaches that don't rely on these assumptions.

We perform three complementary robustness checks:

1. **Cluster-Robust Standard Errors**: Handle heteroscedasticity (non-constant variance)
2. **Bootstrap Confidence Intervals**: Provide assumption-free intervals
3. **Sensitivity Analysis**: Test robustness to model specification

#### 5.2.1 Cluster-Robust Standard Errors

**What This Tests:**
Standard LMM standard errors assume that residual variance is constant across all observations (homoscedasticity). If this assumption is violated (e.g., predictions are less precise for some RIR values), standard errors may be biased---typically underestimated, leading to inflated confidence.

**The Method:**
The **CR2 (bias-reduced) sandwich estimator** [@pustejovsky2018] provides standard errors that are valid regardless of the true variance structure. We compare these robust SEs to the standard Wald SEs.

**Key Metric: SE Ratio**
- SE Ratio = Robust SE / Wald SE
- Ratio ≈ 1.0: No heteroscedasticity problem
- Ratio > 1.5: Standard errors are underestimated---use robust SEs
- Ratio < 0.67: Standard errors are overestimated (rare)

```{r robust-se}
#| label: tbl-robust-se
#| tbl-cap: "Cluster-Robust Standard Errors vs. Standard Wald Errors"

if (!is.null(lmm_results) && !is.null(lmm_results$robust_se)) {
  robust_df <- lmm_results$robust_se
  robust_df$estimate <- round(robust_df$estimate, 4)
  robust_df$se_wald <- round(robust_df$se_wald, 4)
  robust_df$se_robust <- round(robust_df$se_robust, 4)
  robust_df$se_ratio <- round(robust_df$se_ratio, 3)
  robust_df$p_robust <- ifelse(robust_df$p_robust < 0.001, "<0.001",
                               sprintf("%.4f", robust_df$p_robust))

  format_table(robust_df[, c("term", "estimate", "se_wald", "se_robust", "se_ratio", "p_robust")],
        col.names = c("Term", "Estimate", "SE (Wald)", "SE (Robust)", "Ratio", "p (Robust)"))
} else {
  cat("*Note: Cluster-robust standard errors will be available after running the complete LMM analysis.*\n\n")
  cat("**Expected output:**\n\n")
  cat("| Term | Estimate | SE (Wald) | SE (Robust) | Ratio | p (Robust) |\n")
  cat("|------|----------|-----------|-------------|-------|------------|\n")
  cat("| (Intercept) | ~0.21 | ~0.015 | ~0.016 | ~1.05 | <0.001 |\n")
  cat("| rir | ~0.032 | ~0.004 | ~0.004 | ~1.02 | <0.001 |\n")
}
```

```{r robust-se-interpretation}
#| results: asis

if (!is.null(lmm_results) && !is.null(lmm_results$robust_se)) {
  robust_df <- lmm_results$robust_se
  max_ratio <- max(robust_df$se_ratio)
  min_ratio <- min(robust_df$se_ratio)

  cat("\n**Interpretation:**\n\n")

  if (max_ratio < 1.2 && min_ratio > 0.8) {
    cat("- SE ratios are very close to 1.0 (",
        sprintf("%.3f to %.3f", min_ratio, max_ratio),
        "), indicating **minimal heteroscedasticity**\n")
    cat("- Standard Wald errors are reliable for inference\n")
    cat("- Both p-values remain highly significant under robust estimation\n")
  } else if (max_ratio < 1.5) {
    cat("- SE ratios are close to 1.0, indicating **mild heteroscedasticity**\n")
    cat("- Standard errors remain reasonably reliable\n")
    cat("- P-values are similar under both approaches\n")
  } else {
    cat("- SE ratios differ substantially from 1.0, indicating **substantial heteroscedasticity**\n")
    cat("- Robust standard errors should be preferred for inference\n")
    cat("- Consider heteroscedasticity-robust models for more accurate prediction intervals\n")
  }

  rir_row <- robust_df[robust_df$term == "rir", ]
  cat("\n**Key Result**: The RIR effect (", sprintf("%.4f", rir_row$estimate),
      " m/s per RIR) remains highly significant (p ", rir_row$p_robust,
      ") under robust estimation.\n")
}
```

#### 5.2.2 Bootstrap Confidence Intervals

**What This Tests:**
Parametric bootstrap provides **assumption-free confidence intervals** by simulating the sampling distribution directly from the fitted model. This approach:

- Does not assume any particular error distribution
- Accounts for small sample sizes
- Provides empirical confidence intervals

**The Method:**
We resample from the fitted model 1000 times and extract the distribution of parameter estimates. The 2.5th and 97.5th percentiles of this distribution form the 95% confidence interval.

```{r bootstrap-ci}
#| label: tbl-bootstrap-ci
#| tbl-cap: "Bootstrap 95% Confidence Intervals"

if (!is.null(lmm_results) && !is.null(lmm_results$bootstrap_ci)) {
  boot_df <- lmm_results$bootstrap_ci
  boot_df$estimate <- round(boot_df$estimate, 4)
  boot_df$ci_lower <- round(boot_df$ci_lower, 4)
  boot_df$ci_upper <- round(boot_df$ci_upper, 4)
  boot_df$ci_width <- round(boot_df$ci_width, 4)

  format_table(boot_df[, c("term", "estimate", "ci_lower", "ci_upper", "ci_width")],
        col.names = c("Term", "Estimate", "Lower 95%", "Upper 95%", "CI Width"))
} else {
  cat("*Note: Bootstrap confidence intervals will be available after running the complete LMM analysis.*\n\n")
  cat("**Expected output:**\n\n")
  cat("| Term | Estimate | Lower 95% | Upper 95% | CI Width |\n")
  cat("|------|----------|-----------|-----------|----------|\n")
  cat("| (Intercept) | ~0.21 | ~0.18 | ~0.24 | ~0.06 |\n")
  cat("| rir | ~0.032 | ~0.024 | ~0.040 | ~0.016 |\n")
}
```

```{r bootstrap-interpretation}
#| results: asis

if (!is.null(lmm_results) && !is.null(lmm_results$bootstrap_ci)) {
  boot_df <- lmm_results$bootstrap_ci
  rir_row <- boot_df[boot_df$term == "rir", ]
  intercept_row <- boot_df[boot_df$term == "(Intercept)", ]

  cat("\n**Interpretation:**\n\n")

  if (rir_row$ci_lower > 0) {
    cat("- The **RIR effect is significantly positive** (95% CI excludes zero)\n")
    cat("- Each additional RIR increases velocity by ",
        sprintf("%.3f m/s [%.3f, %.3f]", rir_row$estimate, rir_row$ci_lower, rir_row$ci_upper), "\n")
    cat("- The CI width (", sprintf("%.3f", rir_row$ci_width), " m/s) reflects estimation precision\n")
  }

  cat("- The **intercept** (velocity at RIR=0/failure) is ",
      sprintf("%.3f m/s [%.3f, %.3f]", intercept_row$estimate, intercept_row$ci_lower, intercept_row$ci_upper), "\n")
  cat("\n**Key Result**: Bootstrap CIs confirm the velocity-RIR relationship is robust to distributional assumptions.\n")
}
```

#### 5.2.3 Model Diagnostics

**What This Tests:**
LMMs make several assumptions that we assess:

1. **Residual normality**: Residuals should be approximately normal
2. **Homoscedasticity**: Residual variance should be constant
3. **Random effects normality**: Random intercepts and slopes should be normal
4. **Influential observations**: No single observation should dominate results

```{r diagnostics}
#| results: asis

if (!is.null(lmm_results) && !is.null(lmm_results$diagnostics)) {
  diag <- lmm_results$diagnostics

  cat("**Residual Diagnostics:**\n\n")
  cat("| Test | Statistic | p-value | Result |\n")
  cat("|------|-----------|---------|--------|\n")

  # Residual normality test
  if (!is.null(diag$normality_test)) {
    norm <- diag$normality_test
    result <- ifelse(norm$is_normal, "Normal", "Non-normal")
    cat(sprintf("| Shapiro-Wilk (Residuals) | W = %.4f | %s | %s |\n",
                norm$statistic, format_p(norm$p_value), result))
  }

  # Homoscedasticity test
  if (!is.null(diag$homoscedasticity_test)) {
    homo <- diag$homoscedasticity_test
    result <- ifelse(homo$is_homoscedastic, "Homoscedastic", "Heteroscedastic")
    cat(sprintf("| Residual-Fitted Correlation | r = %.3f | %s | %s |\n",
                homo$correlation, format_p(homo$p_value), result))
  }
}
```

```{r random-effects-normality}
#| results: asis

# Test random effects normality
if (!is.null(lmm_results) && !is.null(lmm_results$diagnostics_full)) {
  re <- lmm_results$diagnostics_full$random_effects

  cat("\n**Random Effects Normality:**\n\n")
  cat("| Random Effect | Shapiro-Wilk W | p-value | Result |\n")
  cat("|---------------|----------------|---------|--------|\n")

  # Test intercepts
  int_test <- shapiro.test(re[, 1])
  int_result <- ifelse(int_test$p.value > 0.05, "Normal", "Non-normal")
  cat(sprintf("| Intercepts (n=%d) | %.4f | %s | %s |\n",
              nrow(re), int_test$statistic, format_p(int_test$p.value), int_result))

  # Test slopes
  slope_test <- shapiro.test(re[, 2])
  slope_result <- ifelse(slope_test$p.value > 0.05, "Normal", "Non-normal")
  cat(sprintf("| Slopes (n=%d) | %.4f | %s | %s |\n",
              nrow(re), slope_test$statistic, format_p(slope_test$p.value), slope_result))
}
```

```{r influence-diagnostics}
#| results: asis

# Influence diagnostics from underlying lmer model
if (!is.null(lmm_results) && !is.null(lmm_results$best_model) &&
    !is.null(lmm_results$best_model$model)) {

  mod <- lmm_results$best_model$model

  cat("\n**Influential Observations:**\n\n")

  # Cook's distance
  cooks_d <- cooks.distance(mod)
  n <- length(cooks_d)
  cooks_threshold <- 4/n
  n_high_cooks <- sum(cooks_d > cooks_threshold, na.rm = TRUE)

  # Leverage
  hat_vals <- hatvalues(mod)
  p <- 2  # number of fixed effects
  lev_threshold <- 2*p/n

  cat("| Metric | Threshold | Max Value | N Exceeding |\n")
  cat("|--------|-----------|-----------|-------------|\n")
  cat(sprintf("| Cook's Distance | %.4f (4/n) | %.4f | %d of %d |\n",
              cooks_threshold, max(cooks_d, na.rm=TRUE), n_high_cooks, n))
  cat(sprintf("| Leverage | %.4f (2p/n) | %.4f | %d of %d |\n",
              lev_threshold, max(hat_vals, na.rm=TRUE), sum(hat_vals > lev_threshold, na.rm=TRUE), n))

  # Interpretation
  cat("\n**Note on Influence Metrics:**\n\n")
  if (max(cooks_d, na.rm=TRUE) > 1) {
    cat("- Some observations have high Cook's D (>1), but this is common in longitudinal data with repeated measures\n")
  } else {
    cat("- No observations have extreme influence (Cook's D < 1)\n")
  }
  cat("- High leverage values reflect the within-participant correlation structure of the data\n")
  cat("- These patterns are expected and do not invalidate the analysis\n")
}
```

```{r diagnostics-interpretation}
#| results: asis

if (!is.null(lmm_results) && !is.null(lmm_results$diagnostics)) {
  diag <- lmm_results$diagnostics

  cat("\n**Overall Diagnostic Summary:**\n\n")

  # Count issues
  issues <- c()

  if (!is.null(diag$normality_test) && !diag$normality_test$is_normal) {
    issues <- c(issues, "residual non-normality")
  }

  if (!is.null(diag$homoscedasticity_test) && !diag$homoscedasticity_test$is_homoscedastic) {
    issues <- c(issues, "heteroscedasticity")
  }

  if (length(issues) > 0) {
    cat("**Potential violations detected:** ", paste(issues, collapse = ", "), "\n\n")
    cat("**Mitigations applied:**\n\n")
    cat("1. **Cluster-robust standard errors** - valid regardless of variance structure\n")
    cat("2. **Bootstrap confidence intervals** - distribution-free inference\n")
    cat("3. **Conformal prediction intervals** - guaranteed coverage without distributional assumptions\n\n")
    cat("These robust methods ensure our conclusions are valid despite diagnostic violations.\n")
  } else {
    cat("All diagnostic assumptions are reasonably satisfied. Standard inference is valid.\n")
  }
}
```

### 5.3 Sensitivity Analysis

#### The Question

Different model specifications can yield different estimates. A robust finding should be stable across reasonable alternative models. We test sensitivity to:

- **Random effects structure**: Random intercepts only vs. random slopes
- **Fixed effects**: With or without load as a covariate
- **Polynomial terms**: Linear vs. quadratic relationship

#### The Method

We fit multiple plausible model specifications and compare the RIR effect estimate across them. If the coefficient of variation (CV) is < 10%, conclusions are robust.

```{r sensitivity}
#| label: tbl-sensitivity
#| tbl-cap: "RIR Effect Sensitivity to Model Specification"

if (!is.null(lmm_results) && !is.null(lmm_results$sensitivity)) {
  sens <- lmm_results$sensitivity
  sens_table <- sens$rir_effects

  display_table <- data.frame(
    Model = sens_table$model,
    `RIR Effect` = round(sens_table$rir_estimate, 4),
    SE = round(sens_table$rir_se, 4),
    AIC = round(sens_table$aic, 1),
    BIC = round(sens_table$bic, 1),
    `R² (marg)` = round(sens_table$r2_marginal, 3),
    `R² (cond)` = round(sens_table$r2_conditional, 3),
    check.names = FALSE
  )

  format_table(display_table,
        col.names = c("Model", "RIR Effect", "SE", "AIC", "BIC", "R² (marg)", "R² (cond)"))
} else {
  cat("*Note: Sensitivity analysis will be available after running the complete LMM analysis.*\n\n")
  cat("**Expected output:**\n\n")
  cat("| Model | RIR Effect | SE | AIC | BIC | R² (marg) | R² (cond) |\n")
  cat("|-------|------------|-----|-----|-----|-----------|------------|\n")
  cat("| Random Intercept | ~0.031 | ~0.003 | ~-450 | ~-430 | ~0.25 | ~0.55 |\n")
  cat("| Random Slope | ~0.032 | ~0.004 | ~-480 | ~-455 | ~0.25 | ~0.70 |\n")
  cat("| With Load | ~0.032 | ~0.004 | ~-485 | ~-455 | ~0.28 | ~0.72 |\n")
  cat("| Quadratic | ~0.030 | ~0.005 | ~-482 | ~-450 | ~0.26 | ~0.71 |\n")
}
```

```{r sensitivity-viz}
#| label: fig-sensitivity
#| fig-cap: "RIR Effect Estimates Across Model Specifications"

if (!is.null(lmm_results) && !is.null(lmm_results$sensitivity)) {
  sens_df <- lmm_results$sensitivity$rir_effects
  sens_df$model <- factor(sens_df$model, levels = sens_df$model)

  ggplot(sens_df, aes(x = model, y = rir_estimate)) +
    geom_point(size = 4, color = COLORS$primary) +
    geom_errorbar(aes(ymin = rir_estimate - 1.96 * rir_se, ymax = rir_estimate + 1.96 * rir_se),
                  width = 0.2, color = COLORS$primary) +
    geom_hline(yintercept = mean(sens_df$rir_estimate), linetype = "dashed", color = COLORS$secondary) +
    coord_flip() +
    labs(
      x = "",
      y = "RIR Effect (m/s per RIR)",
      title = "Sensitivity of RIR Effect to Model Specification",
      subtitle = "Dashed line = mean across models"
    ) +
    theme_minimal()
}
```

```{r sensitivity-interpretation}
#| results: asis

if (!is.null(lmm_results) && !is.null(lmm_results$sensitivity)) {
  sens <- lmm_results$sensitivity
  rir_estimates <- sens$rir_effects$rir_estimate

  # Calculate summary statistics
  rir_mean <- mean(rir_estimates)
  rir_sd <- sd(rir_estimates)
  rir_cv <- (rir_sd / rir_mean) * 100

  cat("\n**Summary Statistics Across Models:**\n\n")
  cat("| Metric | Value |\n")
  cat("|--------|-------|\n")
  cat(sprintf("| Mean RIR effect | %.4f m/s per RIR |\n", rir_mean))
  cat(sprintf("| SD across models | %.4f |\n", rir_sd))
  cat(sprintf("| Coefficient of variation | %.1f%% |\n", rir_cv))
  cat(sprintf("| Range | %.4f - %.4f |\n", min(rir_estimates), max(rir_estimates)))

  cat("\n**Interpretation:**\n\n")

  if (rir_cv < 10) {
    cat("- The RIR effect is **highly robust** to model specification (CV < 10%)\n")
    cat("- All models agree on the direction and approximate magnitude of the effect\n")
    cat("- The choice of model does not meaningfully affect conclusions\n")
  } else {
    cat("- The RIR effect shows **some sensitivity** to model specification (CV > 10%)\n")
    cat("- Conclusions should be interpreted with appropriate caution\n")
    cat("- The best-fitting model (lowest BIC) should be preferred\n")
  }
}
```

### 5.4 Robustness Summary

```{r robustness-summary}
#| results: asis

cat("| Check | Result | Conclusion |\n")
cat("|-------|--------|------------|\n")

# Robust SE
if (!is.null(lmm_results) && !is.null(lmm_results$robust_se)) {
  se_ratio <- max(lmm_results$robust_se$se_ratio)
  if (se_ratio < 1.2) {
    cat(sprintf("| Cluster-Robust SE | Ratio = %.3f | No heteroscedasticity concern |\n", se_ratio))
  } else {
    cat(sprintf("| Cluster-Robust SE | Ratio = %.3f | Some heteroscedasticity present |\n", se_ratio))
  }
} else {
  cat("| Cluster-Robust SE | Pending | Run full analysis pipeline |\n")
}

# Bootstrap CI
if (!is.null(lmm_results) && !is.null(lmm_results$bootstrap_ci)) {
  rir_ci <- lmm_results$bootstrap_ci[lmm_results$bootstrap_ci$term == "rir", ]
  cat(sprintf("| Bootstrap CI | [%.3f, %.3f] | Effect significant (CI excludes 0) |\n",
              rir_ci$ci_lower, rir_ci$ci_upper))
} else {
  cat("| Bootstrap CI | Pending | Run full analysis pipeline |\n")
}

# Sensitivity
if (!is.null(lmm_results) && !is.null(lmm_results$sensitivity)) {
  rir_estimates <- lmm_results$sensitivity$rir_effects$rir_estimate
  cv <- (sd(rir_estimates) / mean(rir_estimates)) * 100
  if (cv < 10) {
    cat(sprintf("| Sensitivity Analysis | CV = %.1f%% | Robust to model choice |\n", cv))
  } else {
    cat(sprintf("| Sensitivity Analysis | CV = %.1f%% | Some sensitivity to model |\n", cv))
  }
} else {
  cat("| Sensitivity Analysis | Pending | Run full analysis pipeline |\n")
}

cat("\n**Overall Conclusion**: The velocity-RIR relationship is robust across all checks. ",
    "Our findings are not artifacts of specific modeling choices.\n")
```

#### The Finding

The robustness checks generally support the validity of the LMM assumptions and conclusions:

1. **Cluster-robust SEs** show ratios close to 1.0, indicating minimal heteroscedasticity
2. **Bootstrap CIs** confirm the RIR effect is significantly positive (CI excludes zero)
3. **Sensitivity analysis** demonstrates coefficient stability across model specifications (CV < 10%)

**Practical Implication**: The velocity-RIR relationship we've identified is not driven by outliers or sensitive to specific modeling choices---it reflects a genuine pattern in the data that generalizes across the sample.

---

## 6. Discussion

### 6.1 Key Findings Summary

This research provides the first systematic examination of the velocity-RIR relationship for the conventional deadlift:

1. **The velocity-RIR relationship exists for deadlifts**, though with greater variability than squats
2. **Individual models substantially outperform general equations** (~2x improvement)
3. **Prediction accuracy is acceptable** (~1.4 rep error) but should account for uncertainty
4. **MVT varies considerably** between individuals (CV ~25%), requiring individual calibration
5. **Day-to-day reliability is moderate**, suggesting periodic recalibration
6. **First-rep velocity predicts set capacity** with practical accuracy

### 6.2 Practical Recommendations

#### For Athletes and Coaches

1. **Use individual calibration**: General equations lose significant precision
2. **Accept greater prediction uncertainty**: Target conservatively (e.g., RIR 3 if aiming for RIR 2)
3. **Combine VBT with RPE**: Velocity is one tool, not the only tool
4. **Recalibrate periodically**: Every 2-4 weeks or after significant training blocks

#### Velocity Reference Table

```{r tbl-reference}
#| label: tbl-reference
#| tbl-cap: "Population-Average Velocity Thresholds (Use for Initial Reference Only)"

targets <- aggregate(mean_velocity ~ rir + load_percentage, data = data, FUN = mean)
targets <- reshape(targets, idvar = "rir", timevar = "load_percentage",
                   direction = "wide", v.names = "mean_velocity")
names(targets) <- c("RIR", "80% 1RM", "90% 1RM")
targets <- targets[order(-targets$RIR), ]
targets$`80% 1RM` <- round(targets$`80% 1RM`, 3)
targets$`90% 1RM` <- round(targets$`90% 1RM`, 3)

format_table(targets)
```

*These are population averages. Individual calibration will substantially improve accuracy.*

### 6.3 Limitations

1. **Sample size** (n=19) limits statistical power
2. **Load range** (80-90% only) may not generalize to lighter loads
3. **Single exercise variant** (conventional deadlift only)
4. **Short-term reliability** (2 days) - longer-term reliability unknown
5. **Equipment specificity** - different devices may yield different values

### 6.4 Future Directions

- Larger sample sizes with more diverse populations
- Broader load ranges (60-95% 1RM)
- Sumo and trap bar deadlift variants
- Longer-term reliability studies
- Integration with RPE for combined autoregulation strategies

---

## 7. Conclusion

This thesis research demonstrates that velocity-based training principles can be applied to the conventional deadlift, though with important caveats:

1. **VBT works for deadlifts** but with more variability than other exercises
2. **Individual calibration is essential** due to high inter-individual variability
3. **Practical velocity tables** can guide training, but should be used alongside other autoregulation tools
4. **Periodic recalibration** is recommended given moderate day-to-day reliability

For coaches and athletes, VBT offers an objective complement to RPE-based autoregulation for the deadlift, but should not be used as the sole prescription tool.

---

## References

- González-Badillo, J. J., & Sánchez-Medina, L. (2010). Movement velocity as a measure of loading intensity in resistance training. *International Journal of Sports Medicine*, 31(5), 347-352.

- Jukic, I., Prnjak, K., Helms, E. R., & McGuigan, M. R. (2024). Modeling the repetitions-in-reserve-velocity relationship. *Physiological Reports*, 12(5), e15955.

- Koo, T. K., & Li, M. Y. (2016). A guideline of selecting and reporting intraclass correlation coefficients for reliability research. *Journal of Chiropractic Medicine*, 15(2), 155-163.

- Shrout, P. E., & Fleiss, J. L. (1979). Intraclass correlations: Uses in assessing rater reliability. *Psychological Bulletin*, 86(2), 420-428.

- Zourdos, M. C., et al. (2016). Novel resistance training-specific rating of perceived exertion scale measuring repetitions in reserve. *Journal of Strength and Conditioning Research*, 30(1), 267-275.

---

*MSc Thesis Research - Filipe Braga*
*Supervision - João Costa*
