---
title: "Velocity-Based Training & RIR: A Research Compendium"
subtitle: "Investigating Velocity-Based Training Applications for Powerlifting"
author:
  - name: "João Costa"
    affiliation: "Sport Science Research"
  - name: "Filipe Braga"
    affiliation: "MSc Thesis Research"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# Load results for inline reporting
pelland_results <- readRDS("../data/processed/pelland_replication_results_refactored.rds")
velocity_results <- readRDS("../data/processed/peerj_velocity_replication_results_refactored.rds")
rir_velocity_results <- readRDS("../data/processed/rir_velocity_replication_results.rds")
deadlift_results <- readRDS("../data/processed/deadlift_rir_velocity_results.rds")
lmm_results <- tryCatch(readRDS("../data/processed/deadlift_lmm_results.rds"), error = function(e) NULL)
advanced_results <- tryCatch(readRDS("../data/processed/advanced_velocity_results.rds"), error = function(e) NULL)
```

## About This Research

This research compendium explores fundamental questions in resistance training through velocity-based training (VBT) methods:

> **Can bar velocity objectively measure training proximity to failure?**
>
> **Does this relationship hold for the deadlift, and how can we apply it in practice?**

### Project Structure

This project combines:

1. **Replications** of key published studies (Studies 1-3) to validate methodology
2. **Original research** on the conventional deadlift (Studies 4-6)---part of Filipe Braga's MSc thesis
3. **Practical tools** including velocity stop tables and prediction models

### Research Focus

The original research (Studies 4-6) investigates whether velocity-based training principles, well-established for exercises like the squat and bench press, can be successfully applied to the **conventional deadlift**---an exercise with unique biomechanical characteristics (no stretch-shortening cycle, grip limitations, binary failure patterns).

## Key Concepts

### Repetitions in Reserve (RIR)

RIR measures how close you are to muscular failure:

| RIR | Description | Example |
|-----|-------------|---------|
| 0 | Complete failure | Couldn't complete another rep |
| 1-2 | Near failure | 1-2 reps left before failure |
| 3-4 | Moderate intensity | Noticeable fatigue but not close to failure |
| 5+ | Low intensity | Could do many more reps |

### Velocity-Based Training (VBT)

An objective approach to measuring effort using bar speed:

- **Faster velocity** = More reps remaining
- **Slower velocity** = Closer to failure
- **Velocity loss** = Fatigue accumulation within a set

## Replication Studies (Background)

These replications validate our methodology before applying it to original deadlift research.

### Study 1: Does Training to Failure Matter?

**[Pelland et al. (2024) Replication](01_pelland_meta_regression.qmd)**

A meta-regression examining RIR and training outcomes across strength and hypertrophy studies.

**Key Finding**: Training closer to failure benefits **hypertrophy** but not **strength**.

| Outcome | RIR Effect | Significant? |
|---------|-----------|--------------|
| Strength | ~0 | No |
| Hypertrophy | Negative | Yes |

---

### Study 2: Can Bar Speed Predict Effort?

**[Paulsen et al. (2025) Replication](02_velocity_rir_relationship.qmd)**

Analysis of `r format(nrow(velocity_results$data), big.mark = ",")` observations examining velocity-RIR relationships.

**Key Finding**: Velocity **moderately predicts** perceived effort (r ≈ 0.5), but varies substantially between individuals.

| Metric | Value |
|--------|-------|
| Overall correlation | r = 0.45 |
| Individual range | r = 0.09 to 0.79 |

---

### Study 3: Can We Model the RIR-Velocity Relationship?

**[Jukic et al. (2024) Replication](03_rir_velocity_modeling.qmd)**

Analysis of `r format(nrow(rir_velocity_results$data), big.mark = ",")` observations examining whether individual RIR-velocity models outperform general equations.

**Key Finding**: Individual models explain **~2x more variance** than general models (R² = 0.88 vs 0.50).

| Model Type | Median R² | Implication |
|------------|-----------|-------------|
| General | 0.50 | One-size-fits-all loses precision |
| Individual | 0.88 | Personalization is essential |

---

## Original Deadlift Research (MSc Thesis)

The following studies represent original research investigating the velocity-RIR relationship for the conventional deadlift, forming the core of Filipe Braga's MSc thesis.

### Study 4: Does VBT Work for Deadlifts?

**[Deadlift RIR-Velocity Study](04_deadlift_rir_velocity.qmd)**

Original thesis research applying the same methodology to **conventional deadlift** with `r format(nrow(deadlift_results$data), big.mark = ",")` observations from `r deadlift_results$summary$n_participants` participants.

**Key Finding**: The RIR-velocity relationship holds for deadlifts, but with **lower predictability** than squats.

| Exercise | General R² | Individual R² | Improvement |
|----------|-----------|--------------|-------------|
| Squat (Study 3) | 0.50 | 0.88 | 1.8x |
| Deadlift (Study 4) | `r round(deadlift_results$comparison$general_r2, 2)` | `r round(deadlift_results$comparison$individual_r2, 2)` | `r round(deadlift_results$comparison$improvement_factor, 1)`x |

**Why lower for deadlifts?** Binary failure patterns, grip fatigue, and lack of stretch-shortening cycle.

---

### Study 5: Practical Velocity Tables for Training

**[LMM Analysis & Velocity Stop Tables](05_deadlift_lmm_analysis.qmd)**

Linear Mixed Effects analysis extending Study 4 with practical velocity thresholds for training.

```{r lmm-summary, results='asis'}
if (!is.null(lmm_results)) {
  cat("**Key Finding**: ")
  if (lmm_results$load_importance_result$recommendation == "global") {
    cat("Load percentage does NOT significantly affect the velocity-RIR relationship. A **single global velocity table** is sufficient.\n\n")
  } else {
    cat("Load percentage significantly affects the velocity-RIR relationship. **Load-specific tables** are recommended.\n\n")
  }

  cat("| Table Type | MAE | Use Case |\n")
  cat("|------------|-----|----------|\n")
  cat(sprintf("| General | %.4f m/s | Quick reference, new athletes |\n",
              lmm_results$individual_comparison$global_mae))
  cat(sprintf("| Individual | %.4f m/s | Established athletes with calibration |\n",
              lmm_results$individual_comparison$individual_mae))
  cat(sprintf("\n**Improvement from individualization**: %.1f%%\n",
              lmm_results$individual_comparison$mae_improvement_pct))
} else {
  cat("*Results will be available after running `make analyze-lmm`*\n")
}
```

---

### Study 6: Advanced Velocity-RIR Analyses

**[Advanced Velocity Analyses (H2-H6)](06_advanced_velocity_analyses.qmd)**

Five novel analyses extending the deadlift velocity-RIR research with practical insights.

```{r advanced-summary, results='asis'}
if (!is.null(advanced_results)) {
  mvt <- advanced_results$mvt
  rel <- advanced_results$reliability
  pred <- advanced_results$failure_prediction
  decay <- advanced_results$velocity_decay

  cat("| Analysis | Finding | Practical Implication |\n")
  cat("|----------|---------|----------------------|\n")
  cat(sprintf("| MVT Variability | CV = %.1f%% | Individual calibration essential |\n",
              mvt$population_stats$cv_percent))
  cat(sprintf("| Day-to-Day Reliability | Slope ICC = %.2f | Periodic recalibration needed |\n",
              rel$slope_icc$icc))
  cat(sprintf("| Model Comparison | %.1f%% prefer quadratic | Linear model adequate |\n",
              advanced_results$polynomial_comparison$best_model_summary$pct_quad_best))
  cat(sprintf("| Velocity Decay | %s | %s |\n",
              ifelse(decay$decay_acceleration$accelerating, "Accelerates", "Constant"),
              decay$decay_acceleration$interpretation))
  cat(sprintf("| Failure Prediction | MAE = %.2f reps | First rep velocity useful |\n",
              pred$cv_results$mae))
} else {
  cat("*Results will be available after running `make analyze-advanced`*\n")
}
```

---

## Practical Implications

### For Strength Training

Based on the Pelland meta-regression:

- **Stop 2-3 reps short** of failure on most sets
- Failure training is not necessary for strength gains
- This allows higher training frequency and volume

### For Muscle Building

Based on the Pelland meta-regression:

- **Train closer to failure** (0-2 RIR) for optimal hypertrophy
- The relationship is statistically significant but effect size is modest
- Balance against recovery demands

### For Velocity-Based Training

Based on the Paulsen analysis:

- **Individual calibration is essential** - Group averages don't work for everyone
- **Exercise-specific targets** - Bench press and squat need different thresholds
- **Use as complement** - Velocity + RPE together is better than either alone

### For Deadlift Training

Based on Study 4:

- **VBT works, but with caveats** - Expect ~1.4 rep prediction error (vs ~1 rep for squat)
- **Individual calibration still essential** - 2x improvement over general equations
- **Load-specific thresholds** - Different velocity targets for 80% vs 90% 1RM
- **Consider grip fatigue** - May affect velocity independently of target muscle fatigue

## Summary of Key Findings

Based on all studies in this compendium:

| Question | Finding | Practical Implication |
|----------|---------|----------------------|
| Does VBT work for deadlifts? | Yes, but with more variability than squats | Use VBT as one tool, not the only tool |
| Individual vs general models? | Individual models are ~2x more accurate | Calibrate individually for serious athletes |
| Does load matter? | No significant effect (80% vs 90%) | One velocity table works across loads |
| How often to recalibrate? | Moderate day-to-day reliability | Update every 2-4 weeks |
| Can first rep predict set capacity? | Yes, with ~1-2 rep accuracy | Use for real-time autoregulation |

## Data Sources

| Study | Data | Type |
|-------|------|------|
| Pelland et al. (2024) | [OSF Repository](https://osf.io/7knsj/) | Replication |
| Paulsen et al. (2025) | [PeerJ Supplement](https://peerj.com/articles/19797/) | Replication |
| Jukic et al. (2024) | [OSF Repository](https://osf.io/5ejcp/) | Replication |
| Deadlift Thesis (2024) | Original Data | **Original Research** |

## Methodology

All analyses use:

- **R** for statistical computing
- **lme4** for Linear Mixed Effects Models
- **metafor** for meta-analysis models
- **Quarto** for reproducible reports
- **renv** for dependency management

### Code Architecture

The project follows SOLID principles with R6 classes:

- `R/domain/` - Domain objects (EffectSize, TreatmentGroup)
- `R/calculators/` - Analysis services (RirVelocityModeler, AdvancedVelocityAnalyzer)
- `R/loaders/` - Data loading classes (DeadliftRirDataLoader)
- `scripts/` - Analysis scripts
- `analyses/` - Quarto reports

## References

### Replication Studies

- Robinson ZH, Pelland JC, Zourdos MC, et al. (2024). Exploring the Dose-Response Relationship Between Estimated Resistance Training Proximity to Failure, Strength Gain, and Muscle Hypertrophy. *Sports Medicine*, 54, 2209-2231.

- Paulsen K, Seynnes O, van den Tillaar R. (2025). Exercise type, training load, velocity loss threshold, and sets affect the relationship between lifting velocity and perceived repetitions in reserve. *PeerJ*, 13:e19797.

- Jukic I, Prnjak K, Helms ER, McGuigan MR. (2024). Modeling the repetitions-in-reserve-velocity relationship: A valid method for resistance training monitoring and prescription, and fatigue management. *Experimental Physiology*, 109(2), 193-206.

### Additional References

- González-Badillo JJ, Sánchez-Medina L. (2010). Movement velocity as a measure of loading intensity in resistance training. *International Journal of Sports Medicine*, 31(5), 347-352.

- Zourdos MC, et al. (2016). Novel resistance training–specific rating of perceived exertion scale measuring repetitions in reserve. *Journal of Strength and Conditioning Research*, 30(1), 267-275.

---

*Research conducted by João Costa and Filipe Braga (MSc Thesis)*
