---
title: "RIR & Training Outcomes: A Research Compendium"
subtitle: "Understanding how training intensity affects strength and muscle growth"
author: "Deadlift Study Project"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# Load results for inline reporting
pelland_results <- readRDS("../data/processed/pelland_replication_results_refactored.rds")
velocity_results <- readRDS("../data/processed/peerj_velocity_replication_results_refactored.rds")
rir_velocity_results <- readRDS("../data/processed/rir_velocity_replication_results.rds")
deadlift_results <- readRDS("../data/processed/deadlift_rir_velocity_results.rds")
lmm_results <- tryCatch(readRDS("../data/processed/deadlift_lmm_results.rds"), error = function(e) NULL)
```

## What This Project Investigates

This research compendium explores a fundamental question in resistance training:

> **How close to failure should you train for optimal results?**

We address this through:

1. **Replications** of key published studies
2. **Meta-analysis** of existing literature
3. **Practical guidance** for training decisions

## Key Concepts

### Repetitions in Reserve (RIR)

RIR measures how close you are to muscular failure:

| RIR | Description | Example |
|-----|-------------|---------|
| 0 | Complete failure | Couldn't complete another rep |
| 1-2 | Near failure | 1-2 reps left before failure |
| 3-4 | Moderate intensity | Noticeable fatigue but not close to failure |
| 5+ | Low intensity | Could do many more reps |

### Velocity-Based Training (VBT)

An objective approach to measuring effort using bar speed:

- **Faster velocity** = More reps remaining
- **Slower velocity** = Closer to failure
- **Velocity loss** = Fatigue accumulation within a set

## Completed Analyses

### 1. Does Training to Failure Matter?

**[Pelland et al. (2024) Replication](01_pelland_meta_regression.qmd)**

A meta-regression examining RIR and training outcomes across strength and hypertrophy studies.

**Key Finding**: Training closer to failure benefits **hypertrophy** but not **strength**.

| Outcome | RIR Effect | Significant? |
|---------|-----------|--------------|
| Strength | ~0 | No |
| Hypertrophy | Negative | Yes |

---

### 2. Can Bar Speed Predict Effort?

**[Paulsen et al. (2025) Replication](02_velocity_rir_relationship.qmd)**

Analysis of `r format(nrow(velocity_results$data), big.mark = ",")` observations examining velocity-RIR relationships.

**Key Finding**: Velocity **moderately predicts** perceived effort (r ≈ 0.5), but varies substantially between individuals.

| Metric | Value |
|--------|-------|
| Overall correlation | r = 0.45 |
| Individual range | r = 0.09 to 0.79 |

---

### 3. Can We Model the RIR-Velocity Relationship?

**[Jukic et al. (2024) Replication](03_rir_velocity_modeling.qmd)**

Analysis of `r format(nrow(rir_velocity_results$data), big.mark = ",")` observations examining whether individual RIR-velocity models outperform general equations.

**Key Finding**: Individual models explain **~2x more variance** than general models (R² = 0.88 vs 0.50).

| Model Type | Median R² | Implication |
|------------|-----------|-------------|
| General | 0.50 | One-size-fits-all loses precision |
| Individual | 0.88 | Personalization is essential |

---

### 4. Does This Work for Deadlifts?

**[Deadlift RIR-Velocity Study](04_deadlift_rir_velocity.qmd)**

Original thesis research applying the same methodology to **conventional deadlift** with `r format(nrow(deadlift_results$data), big.mark = ",")` observations from `r deadlift_results$summary$n_participants` participants.

**Key Finding**: The RIR-velocity relationship holds for deadlifts, but with **lower predictability** than squats.

| Exercise | General R² | Individual R² | Improvement |
|----------|-----------|--------------|-------------|
| Squat (Study 3) | 0.50 | 0.88 | 1.8x |
| Deadlift (Study 4) | `r round(deadlift_results$comparison$general_r2, 2)` | `r round(deadlift_results$comparison$individual_r2, 2)` | `r round(deadlift_results$comparison$improvement_factor, 1)`x |

**Why lower for deadlifts?** Binary failure patterns, grip fatigue, and lack of stretch-shortening cycle.

---

### 5. Practical Velocity Tables for Deadlift Training

**[LMM Analysis & Velocity Stop Tables](05_deadlift_lmm_analysis.qmd)**

Linear Mixed Effects analysis extending Study 4 with practical velocity thresholds for training.

```{r lmm-summary, results='asis'}
if (!is.null(lmm_results)) {
  cat("**Key Finding**: ")
  if (lmm_results$load_importance_result$recommendation == "global") {
    cat("Load percentage does NOT significantly affect the velocity-RIR relationship. A **single global velocity table** is sufficient.\n\n")
  } else {
    cat("Load percentage significantly affects the velocity-RIR relationship. **Load-specific tables** are recommended.\n\n")
  }

  cat("| Table Type | MAE | Use Case |\n")
  cat("|------------|-----|----------|\n")
  cat(sprintf("| General | %.4f m/s | Quick reference, new athletes |\n",
              lmm_results$individual_comparison$global_mae))
  cat(sprintf("| Individual | %.4f m/s | Established athletes with calibration |\n",
              lmm_results$individual_comparison$individual_mae))
  cat(sprintf("\n**Improvement from individualization**: %.1f%%\n",
              lmm_results$individual_comparison$mae_improvement_pct))
} else {
  cat("*Results will be available after running `make analyze-lmm`*\n")
}
```

---

## Practical Implications

### For Strength Training

Based on the Pelland meta-regression:

- **Stop 2-3 reps short** of failure on most sets
- Failure training is not necessary for strength gains
- This allows higher training frequency and volume

### For Muscle Building

Based on the Pelland meta-regression:

- **Train closer to failure** (0-2 RIR) for optimal hypertrophy
- The relationship is statistically significant but effect size is modest
- Balance against recovery demands

### For Velocity-Based Training

Based on the Paulsen analysis:

- **Individual calibration is essential** - Group averages don't work for everyone
- **Exercise-specific targets** - Bench press and squat need different thresholds
- **Use as complement** - Velocity + RPE together is better than either alone

### For Deadlift Training

Based on Study 4:

- **VBT works, but with caveats** - Expect ~1.4 rep prediction error (vs ~1 rep for squat)
- **Individual calibration still essential** - 2x improvement over general equations
- **Load-specific thresholds** - Different velocity targets for 80% vs 90% 1RM
- **Consider grip fatigue** - May affect velocity independently of target muscle fatigue

## Data Sources

| Study | Data | Code |
|-------|------|------|
| Pelland et al. (2024) | [OSF Repository](https://osf.io/7knsj/) | Replicated |
| Paulsen et al. (2025) | [PeerJ Supplement](https://peerj.com/articles/19797/) | Replicated |
| Jukic et al. (2024) | [OSF Repository](https://osf.io/5ejcp/) | Replicated |
| Deadlift Thesis (2024) | Original Data | Original Research |

## Methodology

All analyses use:

- **R** for statistical computing
- **metafor** for meta-analysis models
- **Quarto** for reproducible reports
- **renv** for dependency management

### Code Architecture

The project follows SOLID principles with:

- `R/domain/` - Domain objects (EffectSize, TreatmentGroup)
- `R/calculators/` - Effect size calculation services
- `scripts/` - Replication scripts
- `analyses/` - Quarto reports

## Planned Analyses

| Analysis | Description | Status |
|----------|-------------|--------|
| Sensitivity Analysis | Robustness checks | Planned |
| Publication Bias | Bias assessment | Planned |
| ~~Deadlift-Specific~~ | ~~Application to deadlift training~~ | **Completed** (Study 4) |
| ~~LMM & Velocity Tables~~ | ~~Practical velocity thresholds~~ | **Completed** (Study 5) |

## References

- Robinson ZH, Pelland JC, Zourdos MC, et al. (2024). Exploring the Dose-Response Relationship Between Estimated Resistance Training Proximity to Failure, Strength Gain, and Muscle Hypertrophy. *Sports Medicine*, 54, 2209-2231.

- Paulsen K, Seynnes O, van den Tillaar R. (2025). Exercise type, training load, velocity loss threshold, and sets affect the relationship between lifting velocity and perceived repetitions in reserve. *PeerJ*, 13:e19797.

- Jukic I, Prnjak K, Helms ER, McGuigan MR. (2024). Modeling the repetitions-in-reserve-velocity relationship: A valid method for resistance training monitoring and prescription, and fatigue management. *Experimental Physiology*, 109(2), 193-206.
