---
title: "Velocity-Based Training & RIR Research"
subtitle: "Investigating Velocity-Based Training Applications for Powerlifting"
author:
  - name: "Joao Costa"
    affiliation: "Sport Science Research"
  - name: "Filipe Braga"
    affiliation: "MSc Thesis Research"
date: today
---

```{r setup}
#| include: false
source("_common.R")

# Load results for inline reporting
pelland_results <- safe_load_rds("../data/processed/pelland_replication_results.rds")
velocity_results <- safe_load_rds("../data/processed/peerj_velocity_replication_results.rds")
rir_velocity_results <- safe_load_rds("../data/processed/rir_velocity_replication_results.rds")
deadlift_results <- safe_load_rds("../data/processed/deadlift_rir_velocity_results.rds")
lmm_results <- safe_load_rds("../data/processed/deadlift_lmm_results.rds")
advanced_results <- safe_load_rds("../data/processed/advanced_velocity_results.rds")
```

## About This Research

This research compendium explores fundamental questions in resistance training through velocity-based training (VBT) methods:

> **Can bar velocity objectively measure training proximity to failure?**
>
> **Does this relationship hold for the deadlift, and how can we apply it in practice?**

### Project Structure

This project is organized into two main sections:

1. **Background Studies** (Studies 1-3) - Replications of key published research to validate methodology
2. **Deadlift Study (Thesis)** - Original research on the conventional deadlift forming Filipe Braga's MSc thesis

### The Deadlift Challenge

The original research investigates whether velocity-based training principles, well-established for exercises like the squat and bench press, can be successfully applied to the **conventional deadlift**---an exercise with unique biomechanical characteristics:

- No stretch-shortening cycle (concentric-only movement)
- Grip limitations affecting perceived fatigue
- Binary failure patterns (grip release vs gradual deceleration)

---

## Key Concepts

### Repetitions in Reserve (RIR)

RIR measures how close you are to muscular failure:

| RIR | Description | Example |
|-----|-------------|---------|
| 0 | Complete failure | Couldn't complete another rep |
| 1-2 | Near failure | 1-2 reps left before failure |
| 3-4 | Moderate intensity | Noticeable fatigue but not close to failure |
| 5+ | Low intensity | Could do many more reps |

### Velocity-Based Training (VBT)

An objective approach to measuring effort using bar speed:

- **Faster velocity** = More reps remaining
- **Slower velocity** = Closer to failure
- **Velocity loss** = Fatigue accumulation within a set

---

## Background Studies

These replications validate our methodology before applying it to original deadlift research.

### Study 1: Does Training to Failure Matter?

**[Pelland et al. (2024) Replication](01_pelland_meta_regression.qmd)**

A meta-regression examining RIR and training outcomes across strength and hypertrophy studies.

::: {.callout-note}
## Key Finding
Training closer to failure benefits **hypertrophy** but not **strength**.
:::

| Outcome | RIR Effect | Significant? |
|---------|-----------|--------------|
| Strength | ~0 | No |
| Hypertrophy | Negative | Yes |

---

### Study 2: Can Bar Speed Predict Effort?

**[Paulsen et al. (2025) Replication](02_velocity_rir_relationship.qmd)**

Analysis examining velocity-RIR relationships in squat and bench press.

::: {.callout-note}
## Key Finding
Velocity **moderately predicts** perceived effort (r ~ 0.5), but varies substantially between individuals.
:::

| Metric | Value |
|--------|-------|
| Overall correlation | r = 0.45 |
| Individual range | r = 0.09 to 0.79 |

---

### Study 3: General vs Individual Models

**[Jukic et al. (2024) Replication](03_rir_velocity_modeling.qmd)**

Analysis examining whether individual RIR-velocity models outperform general equations.

::: {.callout-note}
## Key Finding
Individual models explain **~2x more variance** than general models (R-squared = 0.88 vs 0.50).
:::

| Model Type | Median R-squared | Implication |
|------------|------------------|-------------|
| General | 0.50 | One-size-fits-all loses precision |
| Individual | 0.88 | Personalization is essential |

---

## Deadlift Study (MSc Thesis)

**[Full Deadlift Study Report](deadlift_study.qmd)**

This comprehensive report presents original thesis research investigating the velocity-RIR relationship for the **conventional deadlift**. It combines:

- **Exploratory analysis** of the deadlift velocity-RIR relationship
- **Linear Mixed Effects Models** with nested random effects
- **Practical velocity stop tables** for training prescription
- **Advanced analyses** including reliability, model comparison, and failure prediction

### Key Findings Summary

```{r thesis-summary}
#| results: asis

if (!is.null(deadlift_results)) {
  cat("| Metric | Deadlift | Squat (Reference) |\n")
  cat("|--------|----------|-------------------|\n")
  cat(sprintf("| General R-squared | %.2f | 0.50 |\n", deadlift_results$comparison$general_r2))
  cat(sprintf("| Individual R-squared | %.2f | 0.88 |\n", deadlift_results$comparison$individual_r2))
  cat(sprintf("| Improvement Factor | %.1fx | 1.8x |\n", deadlift_results$comparison$improvement_factor))
  cat(sprintf("| Participants | %d | 19 |\n", deadlift_results$summary$n_participants))
} else {
  cat("*Results will be available after running the analysis pipeline*\n")
}
```

### Why Lower for Deadlifts?

The deadlift presents unique challenges for velocity-based training:

1. **Binary failure** - Grip often fails suddenly rather than gradual velocity decline
2. **No eccentric phase** - Each rep starts from a dead stop (no stretch-shortening cycle)
3. **Grip fatigue** - May affect velocity independently of target muscle fatigue

---

## Summary of Key Findings

Based on all studies in this compendium:

| Question | Finding | Practical Implication |
|----------|---------|----------------------|
| Does VBT work for deadlifts? | Yes, but with more variability than squats | Use VBT as one tool, not the only tool |
| Individual vs general models? | Individual models are ~2x more accurate | Calibrate individually for serious athletes |
| Does load matter? | No significant effect (80% vs 90%) | One velocity table works across loads |
| How often to recalibrate? | Moderate day-to-day reliability | Update every 2-4 weeks |
| Can first rep predict set capacity? | Yes, with ~1-2 rep accuracy | Use for real-time autoregulation |

---

## Practical Implications

### For Strength Training

Based on the Pelland meta-regression:

- **Stop 2-3 reps short** of failure on most sets
- Failure training is not necessary for strength gains
- This allows higher training frequency and volume

### For Muscle Building

Based on the Pelland meta-regression:

- **Train closer to failure** (0-2 RIR) for optimal hypertrophy
- The relationship is statistically significant but effect size is modest
- Balance against recovery demands

### For Velocity-Based Training

Based on the Paulsen analysis:

- **Individual calibration is essential** - Group averages don't work for everyone
- **Exercise-specific targets** - Different exercises need different thresholds
- **Use as complement** - Velocity + RPE together is better than either alone

### For Deadlift Training

Based on the thesis research:

- **VBT works, but with caveats** - Expect ~1.4 rep prediction error (vs ~1 rep for squat)
- **Individual calibration still essential** - 2x improvement over general equations
- **Consider grip fatigue** - May affect velocity independently of target muscle fatigue

---

## Data Sources

| Study | Data | Type |
|-------|------|------|
| Pelland et al. (2024) | [OSF Repository](https://osf.io/7knsj/) | Replication |
| Paulsen et al. (2025) | [PeerJ Supplement](https://peerj.com/articles/19797/) | Replication |
| Jukic et al. (2024) | [OSF Repository](https://osf.io/5ejcp/) | Replication |
| Deadlift Thesis (2024) | Original Data | **Original Research** |

## Methodology

All analyses use:

- **R** for statistical computing
- **lme4** for Linear Mixed Effects Models
- **metafor** for meta-analysis models
- **Quarto** for reproducible reports
- **renv** for dependency management

### Code Architecture

The project follows SOLID principles with R6 classes:

- `R/domain/` - Domain objects (EffectSize, TreatmentGroup)
- `R/calculators/` - Analysis services (RirVelocityModeler, AdvancedVelocityAnalyzer)
- `R/loaders/` - Data loading classes (DeadliftRirDataLoader)
- `scripts/` - Analysis scripts
- `analyses/` - Quarto reports

---

*Research conducted by Joao Costa and Filipe Braga (MSc Thesis)*
